---
title: "Agricultural Variables — Continuous Distributions"
subtitle: "Quartile and Quintile Analysis | São Paulo state municipalities"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

## 1. Setup

```{r}
library(arrow)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

source(file.path(rprojroot::find_rstudio_root_file(), "config.R"))
```

---

## 2. Load and prepare municipality-level base

PAM and CAGED are annual — averaged across available years. Fixed covariates (Censo Agro 2017, Urban/Rural 2022) take the first non-NA value since they repeat across all years.

```{r}
base_full <- read_parquet(
  file.path(project_root, "resultados/base_consolidada_sp_2014_2024.parquet")
) |>
  mutate(cod_ibge = as.character(cod_ibge))

base_mun <- base_full |>
  select(cod_ibge, nome_municipio,
         pam_area_colhida_ha, pam_valor_prod_mil_reais,
         censo_pct_uso_agrotox, censo_valor_agrotox_mil,
         censo_uso_total_estab,
         caged_admissoes_agro,
         pct_rural_2022) |>
  group_by(cod_ibge, nome_municipio) |>
  summarise(
    pam_area_colhida_ha_media      = mean(pam_area_colhida_ha,        na.rm = TRUE),
    pam_valor_prod_mil_reais_media = mean(pam_valor_prod_mil_reais,   na.rm = TRUE),
    caged_admissoes_agro_media     = mean(caged_admissoes_agro,       na.rm = TRUE),
    censo_pct_uso_agrotox          = first(na.omit(censo_pct_uso_agrotox)),
    censo_valor_agrotox_mil        = first(na.omit(censo_valor_agrotox_mil)),
    censo_uso_total_estab          = first(na.omit(censo_uso_total_estab)),
    pct_rural_2022                 = first(na.omit(pct_rural_2022)),
    .groups = "drop"
  ) |>
  mutate(
    agrotox_por_hectare = ifelse(
      pam_area_colhida_ha_media > 0,
      censo_valor_agrotox_mil / pam_area_colhida_ha_media,
      NA_real_
    )
  )

cat("Municipalities:", nrow(base_mun))
```

---

## 3. Helper functions

```{r}
# Numeric summary
summarise_continuous <- function(df, var) {
  x <- df[[var]]
  x_valid <- x[!is.na(x)]
  tibble(
    Variable  = var,
    N         = length(x_valid),
    N_missing = sum(is.na(x)),
    Min       = round(min(x_valid), 2),
    Mean      = round(mean(x_valid), 2),
    Median    = round(median(x_valid), 2),
    Max       = round(max(x_valid), 2),
    SD        = round(sd(x_valid), 2)
  )
}

# Quantile frequency table
make_quantile_table <- function(df, var, n_breaks = 5) {
  x <- df[[var]]
  x_valid <- x[!is.na(x)]
  probs  <- seq(0, 1, length.out = n_breaks + 1)
  breaks <- unique(quantile(x_valid, probs = probs))

  if (length(breaks) < 3) return(tibble(Note = "Not enough distinct values"))

  group_label <- if (n_breaks == 4) "Quartile" else "Quintile"
  labels <- paste0(group_label, " ", seq_len(length(breaks) - 1))
  grp <- cut(x_valid, breaks = breaks, labels = labels, include.lowest = TRUE)

  as.data.frame(table(grp)) |>
    rename(Group = grp, N = Freq) |>
    mutate(
      `%`   = round(100 * N / sum(N), 1),
      Range = paste0("[",
                     round(breaks[-length(breaks)], 2),
                     " - ",
                     round(breaks[-1], 2),
                     "]")
    ) |>
    select(Group, Range, N, `%`)
}

# Histogram helper
plot_hist <- function(df, var, label) {
  x <- df[[var]]
  x <- x[!is.na(x)]
  ggplot(data.frame(x = x), aes(x)) +
    geom_histogram(bins = 30, fill = "steelblue", colour = "white", alpha = 0.8) +
    labs(title = label, x = NULL, y = "Count") +
    theme_minimal()
}
```

---

## 4. PAM — Mean harvested area (ha)

```{r}
summarise_continuous(base_mun, "pam_area_colhida_ha_media") |> kable()
make_quantile_table(base_mun, "pam_area_colhida_ha_media", n_breaks = 5) |> kable()
plot_hist(base_mun, "pam_area_colhida_ha_media", "PAM — Mean harvested area (ha)")
```

---

## 5. Censo Agro — % establishments using pesticides

```{r}
summarise_continuous(base_mun, "censo_pct_uso_agrotox") |> kable()
make_quantile_table(base_mun, "censo_pct_uso_agrotox", n_breaks = 5) |> kable()
plot_hist(base_mun, "censo_pct_uso_agrotox", "Censo Agro — % establishments using pesticides")
```

---

## 6. Censo Agro — Pesticide spending (thousand BRL)

```{r}
summarise_continuous(base_mun, "censo_valor_agrotox_mil") |> kable()
make_quantile_table(base_mun, "censo_valor_agrotox_mil", n_breaks = 5) |> kable()
plot_hist(base_mun, "censo_valor_agrotox_mil", "Censo Agro — Pesticide spending (thousand BRL)")
```

---

## 7. PAM — Mean agricultural production value (thousand BRL)

```{r}
summarise_continuous(base_mun, "pam_valor_prod_mil_reais_media") |> kable()
make_quantile_table(base_mun, "pam_valor_prod_mil_reais_media", n_breaks = 5) |> kable()
plot_hist(base_mun, "pam_valor_prod_mil_reais_media", "PAM — Mean production value (thousand BRL)")
```

---

## 8. CAGED — Mean formal agricultural admissions

```{r}
summarise_continuous(base_mun, "caged_admissoes_agro_media") |> kable()
make_quantile_table(base_mun, "caged_admissoes_agro_media", n_breaks = 5) |> kable()
plot_hist(base_mun, "caged_admissoes_agro_media", "CAGED — Mean formal agricultural admissions")
```

> **Note:** 17 months in 2014–2019 are missing from CAGED due to corrupt source files.
> Annual means for those years are underestimates.

---

## 9. Urban/Rural — % rural population (Census 2022)

```{r}
summarise_continuous(base_mun, "pct_rural_2022") |> kable()
make_quantile_table(base_mun, "pct_rural_2022", n_breaks = 5) |> kable()
plot_hist(base_mun, "pct_rural_2022", "Urban/Rural — % rural population (Census 2022)")
```

---

## 10. Derived — Pesticide spending per harvested hectare

```{r}
summarise_continuous(base_mun, "agrotox_por_hectare") |> kable()
make_quantile_table(base_mun, "agrotox_por_hectare", n_breaks = 5) |> kable()
plot_hist(base_mun, "agrotox_por_hectare", "Derived — Pesticide spending per harvested hectare")
```

---

## 11. Save municipality base with quantile group columns

```{r}
assign_quantile_group <- function(x, n_breaks = 5) {
  probs  <- seq(0, 1, length.out = n_breaks + 1)
  breaks <- unique(quantile(x, probs = probs, na.rm = TRUE))
  cut(x, breaks = breaks,
      labels = paste0("Q", seq_len(length(breaks) - 1)),
      include.lowest = TRUE)
}

base_mun_groups <- base_mun |>
  mutate(
    q5_area_colhida    = assign_quantile_group(pam_area_colhida_ha_media,      5),
    q5_pct_agrotox     = assign_quantile_group(censo_pct_uso_agrotox,          5),
    q5_valor_agrotox   = assign_quantile_group(censo_valor_agrotox_mil,        5),
    q5_valor_prod      = assign_quantile_group(pam_valor_prod_mil_reais_media, 5),
    q5_caged_admissoes = assign_quantile_group(caged_admissoes_agro_media,     5),
    q4_pct_rural       = assign_quantile_group(pct_rural_2022,                 4),
    q5_agrotox_hectare = assign_quantile_group(agrotox_por_hectare,            5)
  )

write_parquet(
  base_mun_groups,
  file.path(project_root, "resultados/agro_municipios_grupos_quantis.parquet")
)

cat("Saved: resultados/agro_municipios_grupos_quantis.parquet\n")
glimpse(base_mun_groups)
```
