---
title: "Population Pyramids — São Paulo Municipalities"
subtitle: "Showcase using populacao_sp_municipio_ano.parquet"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

## 1. Setup

```{r}
library(arrow)
library(dplyr)
library(tidyr)
library(ggplot2)

source(file.path(rprojroot::find_rstudio_root_file(), "config.R"))
```

---

## 2. Load population data

The source file has one row per municipality × year × sex × single-year age.

```{r}
pop_raw <- read_parquet(
  file.path(project_root, "resultados/contextual/populacao_sp_municipio_ano.parquet")
)

glimpse(pop_raw)
```

---

## 3. Build age groups (5-year bands)

Single-year age is too granular for a readable pyramid. Group into standard 5-year bands, with an open-ended "80+" top band.

```{r}
pop_grouped <- pop_raw |>
  mutate(
    faixa_etaria = case_when(
      idade < 5   ~ "0-4",
      idade < 10  ~ "5-9",
      idade < 15  ~ "10-14",
      idade < 20  ~ "15-19",
      idade < 25  ~ "20-24",
      idade < 30  ~ "25-29",
      idade < 35  ~ "30-34",
      idade < 40  ~ "35-39",
      idade < 45  ~ "40-44",
      idade < 50  ~ "45-49",
      idade < 55  ~ "50-54",
      idade < 60  ~ "55-59",
      idade < 65  ~ "60-64",
      idade < 70  ~ "65-69",
      idade < 75  ~ "70-74",
      idade < 80  ~ "75-79",
      TRUE        ~ "80+"
    ),
    faixa_etaria = factor(faixa_etaria, levels = c(
      "0-4","5-9","10-14","15-19","20-24","25-29","30-34","35-39","40-44",
      "45-49","50-54","55-59","60-64","65-69","70-74","75-79","80+"
    ))
  ) |>
  group_by(cod_ibge, ano, sexo, faixa_etaria) |>
  summarise(populacao = sum(populacao, na.rm = TRUE), .groups = "drop")
```

---

## 4. Pyramid function (single municipality × year)

Male population is plotted as negative values so it extends left, female extends right — the standard pyramid convention.

```{r}
plot_pyramid <- function(df, municipio_cod, ano_ref, municipio_nome = NULL) {

  d <- df |>
    filter(cod_ibge == municipio_cod, ano == ano_ref) |>
    mutate(
      populacao_plot = if_else(sexo == "Masculino", -populacao, populacao)
    )

  titulo <- if (!is.null(municipio_nome)) {
    paste0(municipio_nome, " — ", ano_ref)
  } else {
    paste0("Municipality ", municipio_cod, " — ", ano_ref)
  }

  max_val <- max(d$populacao, na.rm = TRUE)

  ggplot(d, aes(x = faixa_etaria, y = populacao_plot, fill = sexo)) +
    geom_col(width = 0.85) +
    coord_flip() +
    scale_y_continuous(
      labels = \(x) scales::comma(abs(x)),
      limits = c(-max_val * 1.1, max_val * 1.1)
    ) +
    scale_fill_manual(values = c("Masculino" = "#1C7293", "Feminino" = "#C0392B")) +
    labs(
      title = titulo,
      x = "Age group", y = "Population", fill = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "top",
      panel.grid.minor = element_blank()
    )
}
```

---

## 5. Example — São Paulo capital (cod_ibge 355030), 2024

```{r}
plot_pyramid(pop_grouped, "355030", 2024, "São Paulo (capital)")
```

---

## 6. Example — comparing two years for the same municipality

Useful to visualize population aging over time.

```{r}
plot_pyramid(pop_grouped, "355030", 2014, "São Paulo (capital)")
```

```{r}
plot_pyramid(pop_grouped, "355030", 2024, "São Paulo (capital)")
```

---

## 7. Faceted pyramid — multiple municipalities, same year

Compare the population structure of a few municipalities side by side.

```{r}
municipios_exemplo <- c("355030", "350010", "354880")  # adjust codes as needed
ano_exemplo <- 2024

pop_grouped |>
  filter(cod_ibge %in% municipios_exemplo, ano == ano_exemplo) |>
  mutate(populacao_plot = if_else(sexo == "Masculino", -populacao, populacao)) |>
  ggplot(aes(x = faixa_etaria, y = populacao_plot, fill = sexo)) +
  geom_col(width = 0.85) +
  coord_flip() +
  facet_wrap(~cod_ibge, scales = "free_x") +
  scale_y_continuous(labels = \(x) scales::comma(abs(x))) +
  scale_fill_manual(values = c("Masculino" = "#1C7293", "Feminino" = "#C0392B")) +
  labs(title = paste("Population structure —", ano_exemplo),
       x = "Age group", y = "Population", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

---

## 8. Notes

- Replace the municipality codes in Sections 5–7 with any valid `cod_ibge` from the dataset to explore other municipalities.
- For rate or share-based pyramids (% of total population per band, rather than absolute counts), divide `populacao` by the municipality's `pop_total` for that year before plotting — useful when comparing municipalities of very different sizes.
