---
title: "State of São Paulo ,Brazil population Descriptive_analysis"
author: "Abirami Somasundaram"
date: "2026-06-08"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown Notes

1. This is a .rmd document. 
2. This is the analaysis of 2 datasets. 
  a. ** File** - 'base_consolidada_sp_2014_2024.parquet'
  
  | # | Variable | Type | Description |
|---|---|---|---|
| 4 | `pop_total` | integer | Total estimated population |
| 5 | `pop_masculino` | integer | Male population |
| 6 | `pop_feminino` | integer | Female population |
| 7 | `pop_0_14` | integer | Population aged 0–14 |
| 8 | `pop_15_64` | integer | Population aged 15–64 |
| 9 | `pop_65plus` | integer | Population aged 65 and older |

Source: `resultados/contextual/populacao_sp_municipio_ano.parquet`

b. populacao_sp_municipio_ano.parquet 
 
  | Variable | Type | Description |
|---|---|---|
| `cod_ibge` | string | 6-digit IBGE municipality code |
| `ano` | integer | Reference year |
| `sexo` | string | `"Masculino"` or `"Feminino"` |
| `idade` | integer | Single year of age |
| `populacao` | integer | Estimated population count |

**Source:** IBGE intercensal population estimates via DATASUS

3. The Descriptive analysis has the following: 
  - Creating quartiles for the population variables and calculating mean, median, range and count. 
  - The children's age is diveded in 0-3, 4-7,8-10 and 10-14.
  -Then quartiles were created for each age group grouped by year and calculated mean, median, range and count. 
  - Calculating mean, median and range for for population age group 0-14 by sex 
  - Proportion of Child Population by Age Group Over Time
  -  Percent change in proportion
  -Plots for propotion variation year wise and sex wise. 
  

```{r include=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(arrow)
library(purrr)
library(scales)
library(tidyverse)
library(stringr)
library(ggh4x)
library(tidyverse)

```

Load data
```{r include=FALSE}
pop_data <- read_parquet("/Users/abiramisomasundaram/Desktop/Brazil pesticides/Data/Dados Version 2/Dados/resultados/base_consolidada_sp_2014_2024.parquet") %>%
select(
    "cod_ibge",
  "nome_municipio",
  "ano",
  "pop_total",
  "pop_masculino",
  "pop_feminino",
  "pop_0_14",
  "pop_15_64",
  "pop_65plus" )

```

# 1. Age descriptive analysis

```{r include=TRUE}
pop_data <- pop_data %>% 
mutate( 
pct_0_14 = if_else( 
pop_total > 0, 
pop_0_14 / pop_total * 100, 
NA_real_ 
), 
pct_15_64 = if_else( 
pop_total > 0, 
pop_15_64 / pop_total * 100, 
NA_real_ 
), 
pct_65_plus = if_else( 
  pop_total > 0, 
  pop_65plus / pop_total * 100, 
NA_real_ 
) 
)
head(pop_data)
```

# 2. Descriptive Statistics of Raw Counts - Quartiles grouped by year

This section summarizes the raw population counts using standard descriptive statistics. These include measures such as:

Quartiles
Mean population count  
Median population count  
Minimum and maximum values  
Range  


```{r}
pop_vars <- c( 
  "pop_total",
  "pop_masculino",
  "pop_feminino",
  "pop_0_14",
  "pop_15_64",
  "pop_65plus" )

quartile_summary_by_year <- pop_data %>%
  pivot_longer(
    cols = all_of(pop_vars),
    names_to = "variable",
    values_to = "value"
  ) %>%
  filter(!is.na(value)) %>%
  group_by(ano, variable) %>%
  mutate(
    Quartile = paste0("Q", ntile(value, 4))
  ) %>%
  group_by(ano, variable, Quartile) %>%
  summarise(
    n = n(),
    Mean = mean(value, na.rm = TRUE),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    `%` = round(100 * n / sum(n), 2),
    Range = paste0(Min, " - ", Max)
  ) %>%
  ungroup() %>%
  select(
    ano,
    variable,
    Quartile,
    n,
    `%`,
    Mean,
    Median,
    Range
  ) %>%
  arrange(ano, variable, Quartile)

head(quartile_summary_by_year) 

```

# 3. Quartile by spliting the variable pop_0-14

This section summarizes the raw population counts for childeren of age 0-14  using standard descriptive statistics. These include measures such as:

Quartiles
Mean population count  
Median population count  
Minimum and maximum values  
Range  


```{r}
child_data <- read_parquet(
  "/Users/abiramisomasundaram/Desktop/Brazil pesticides/Data/Dados Version 2/Dados/resultados/contextual/populacao_sp_municipio_ano.parquet"
)

child_age_quartiles <- child_data %>%
  filter(idade >= 0, idade <= 14) %>%
  mutate(
    age_group = case_when(
      idade >= 0  & idade <= 3  ~ "0-3",
      idade >= 4  & idade <= 7  ~ "4-7",
      idade >= 8  & idade <= 10 ~ "8-10",
      idade >= 11 & idade <= 14 ~ "11-14",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_group)) %>%
  group_by(ano, cod_ibge, age_group) %>%
  summarise(
    population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(ano, age_group) %>%
  mutate(
    Quartile = paste0("Q", ntile(population, 4))
  ) %>%
  ungroup()

head(child_age_quartiles)
  
```


# 4. Calculating mean, median and range for for population age group 0-14 by year


```{r}

child_quartile_summary <- child_age_quartiles %>%
  group_by(ano, age_group, Quartile) %>%
  summarise(
    n = n(),
    Mean = mean(population, na.rm = TRUE),
    Median = median(population, na.rm = TRUE),
    Min = min(population, na.rm = TRUE),
    Max = max(population, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    percent = round(100 * n / sum(n), 2),
    Range = paste0(Min, " - ", Max)
  ) %>%
  ungroup() %>%
  select(
    ano,
    age_group,
    Quartile,
    n,
    percent,
    Mean,
    Median,
    Range
  ) %>%
  arrange(ano, age_group, Quartile)

head(child_quartile_summary)


```



# 5.  Calculating mean, median and range for for population age group 0-14 by sex

```{r}
child_age_quartiles_sex <- child_data %>%
  filter(idade >= 0, idade <= 14) %>%
  mutate(
    age_group = case_when(
      idade >= 0  & idade <= 3  ~ "0-3",
      idade >= 4  & idade <= 7  ~ "4-7",
      idade >= 8  & idade <= 10 ~ "8-10",
      idade >= 11 & idade <= 14 ~ "11-14",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_group)) %>%
  group_by(ano, cod_ibge, sexo, age_group) %>%
  summarise(
    population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(ano, sexo, age_group) %>%
  mutate(
    Quartile = paste0("Q", ntile(population, 4))
  ) %>%
  ungroup()

child_quartile_summary_sex <- child_age_quartiles_sex %>%
  group_by(ano, sexo, age_group, Quartile) %>%
  summarise(
    n = n(),
    Mean = mean(population, na.rm = TRUE),
    Median = median(population, na.rm = TRUE),
    Min = min(population, na.rm = TRUE),
    Max = max(population, na.rm = TRUE),
    .groups = "drop_last"
  ) %>%
  mutate(
    percent = round(100 * n / sum(n), 2),
    Range = paste0(Min, " - ", Max)
  ) %>%
  ungroup() %>%
  arrange(ano, sexo, age_group, Quartile)

head(child_quartile_summary_sex)

```

#6. Create age groups and calculation of  yearly proportions(%)

```{r}
child_age_yearly <- child_data %>%
  filter(idade >= 0, idade <= 14) %>%
  mutate(
    age_group = case_when(
      idade >= 0  & idade <= 3  ~ "0-3",
      idade >= 4  & idade <= 7  ~ "4-7",
      idade >= 8  & idade <= 10 ~ "8-10",
      idade >= 11 & idade <= 14 ~ "11-14",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_group)) %>%
  group_by(ano, age_group) %>%
  summarise(
    total_population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(ano) %>%
  mutate(
    total_child_population = sum(total_population, na.rm = TRUE),
    proportion = total_population / total_child_population,
    proportion_percent = 100 * proportion
  ) %>%
  ungroup() %>%
  arrange(age_group, ano)

head(child_age_yearly)

```

# 7. Percent change in proportion

```{r}

child_age_variation <- child_age_yearly %>%
  group_by(age_group) %>%
  arrange(ano, .by_group = TRUE) %>%
  mutate(
    previous_proportion_percent = lag(proportion_percent),
    
    percentage_point_change = proportion_percent - previous_proportion_percent,
    
    percent_change_in_proportion = 100 * 
      (proportion_percent - previous_proportion_percent) / previous_proportion_percent
  ) %>%
  ungroup()

head(child_age_variation)
```


# 8. Proportion of Child Population by Age Group Over Time
This shows the percentage share of each age group each year.
```{r}
ggplot(
  child_age_variation,
  aes(
    x = ano,
    y = proportion_percent,
    color = age_group,
    group = age_group
  )
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Proportion of Child Population by Age Group Over Time",
    x = "Year",
    y = "Percentage of population age 0-14",
    color = "Age group"
  ) +
  theme_minimal()
```

#8. Proportional variation year-wise

```{r}
ggplot(
  child_age_variation,
  aes(
    x = ano,
    y = percentage_point_change,
    color = age_group,
    group = age_group
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Year-wise Change in Population Proportion by Age Group",
    x = "Year",
    y = "Percentage point change from previous year",
    color = "Age group"
  ) +
  theme_minimal()
```


```{r}
child_age_yearly_sex <- child_data %>%
  filter(idade >= 0, idade <= 14) %>%
  mutate(
    age_group = case_when(
      idade >= 0  & idade <= 3  ~ "0-3",
      idade >= 4  & idade <= 7  ~ "4-7",
      idade >= 8  & idade <= 10 ~ "8-10",
      idade >= 11 & idade <= 14 ~ "11-14",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_group)) %>%
  group_by(ano, sexo, age_group) %>%
  summarise(
    total_population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(ano, sexo) %>%
  mutate(
    total_child_population = sum(total_population, na.rm = TRUE),
    proportion_percent = 100 * total_population / total_child_population
  ) %>%
  ungroup() %>%
  arrange(sexo, age_group, ano)

child_age_variation_sex <- child_age_yearly_sex %>%
  group_by(sexo, age_group) %>%
  arrange(ano, .by_group = TRUE) %>%
  mutate(
    previous_proportion_percent = lag(proportion_percent),
    percentage_point_change = proportion_percent - previous_proportion_percent,
    percent_change_in_proportion = 100 *
      (proportion_percent - previous_proportion_percent) / previous_proportion_percent
  ) %>%
  ungroup()

#Plot 
ggplot(
  child_age_variation_sex,
  aes(
    x = ano,
    y = percentage_point_change,
    color = age_group,
    group = age_group
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ sexo) +
  labs(
    title = "Year-wise Change in Child Population Proportion by Age Group and Sex",
    x = "Year",
    y = "Percentage point change from previous year",
    color = "Age group"
  ) +
  theme_minimal()
```








