---
title: "State of São Paulo ,Brazil population Descriptive_analysis"
author: "Abirami Somasundaram"
date: "2026-06-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown Notes

1. This is a .rmd document. 
2. This is the analaysis of 2 datasets. 
  a. ** File** - 'base_consolidada_sp_2014_2024.parquet'
  
  | # | Variable | Type | Description |
|---|---|---|---|
| 4 | `pop_total` | integer | Total estimated population |
| 5 | `pop_masculino` | integer | Male population |
| 6 | `pop_feminino` | integer | Female population |
| 7 | `pop_0_14` | integer | Population aged 0–14 |
| 8 | `pop_15_64` | integer | Population aged 15–64 |
| 9 | `pop_65plus` | integer | Population aged 65 and older |

Source: `resultados/contextual/populacao_sp_municipio_ano.parquet`

b. populacao_sp_municipio_ano.parquet 
 
  | Variable | Type | Description |
|---|---|---|
| `cod_ibge` | string | 6-digit IBGE municipality code |
| `ano` | integer | Reference year |
| `sexo` | string | `"Masculino"` or `"Feminino"` |
| `idade` | integer | Single year of age |
| `populacao` | integer | Estimated population count |

**Source:** IBGE intercensal population estimates via DATASUS

3. The Descriptive analysis has the following: 
  - Creating quartiles for the population variables and calculating mean, median, range and count. 
  - The children's age is diveded in 0-3, 4-7,8-10 and 10-14.
  -Then quartiles were created for each age group grouped by year and calculated mean, median, range and count. 
  - Calculating mean, median and range for for population age group 0-14 by sex 
  - Proportion of Child Population by Age Group Over Time
  -  Percent change in proportion
  -Plots for propotion variation year wise and sex wise. 
  


```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(scales)
library(sf)
library(geobr)
library(knitr)
library(viridis)
library(arrow)
```


Load data
```{r}
pop <- read_parquet("/Users/abiramisomasundaram/Desktop/Brazil pesticides/Data/Dados Version 2/Dados/resultados/contextual/populacao_sp_municipio_ano.parquet") 

pop <- pop %>%
  clean_names() %>%
  mutate(
    cod_ibge = as.character(cod_ibge),
    cod_ibge = str_pad(cod_ibge, width = 6, side = "left", pad = "0"),
    ano = as.integer(ano),
    sexo = as.character(sexo),
    idade = as.integer(idade),
    populacao = as.numeric(populacao)
  )

glimpse(pop)

```



# 1. Missingness of Whole database

```{r missingness-whole-database}
missingness_table <- pop %>%
  summarise(
    across(
      everything(),
      list(
        n_missing = ~sum(is.na(.x)),
        pct_missing = ~mean(is.na(.x)) * 100
      )
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", ".value"),
    names_pattern = "(.+)_(n_missing|pct_missing)"
  ) %>%
  arrange(desc(pct_missing)) %>%
  mutate(
    pct_missing = round(pct_missing, 2)
  )

kable(
  missingness_table,
  caption = "Missingness for the full population database"
)
```

```{r complete-cases}
complete_case_summary <- pop %>%
  summarise(
    total_rows = n(),
    complete_rows = sum(complete.cases(.)),
    incomplete_rows = sum(!complete.cases(.)),
    pct_incomplete = round(mean(!complete.cases(.)) * 100, 2)
  )

kable(
  complete_case_summary,
  caption = "Complete and incomplete rows in the full database"
)
```


# 3. Restrict to children aged 0-14

```{r child-age-groups}
pop_child <- pop %>%
  filter(idade >= 0, idade <= 14) %>%
  mutate(
    age_group = case_when(
      idade >= 0  & idade <= 3  ~ "0-3",
      idade >= 4  & idade <= 7  ~ "4-7",
      idade >= 8  & idade <= 10 ~ "8-10",
      idade >= 11 & idade <= 14 ~ "11-14",
      TRUE ~ NA_character_
    ),
    age_group = factor(
      age_group,
      levels = c("0-3", "4-7", "8-10", "11-14")
    )
  )

glimpse(pop_child)
```


# 4. Frequency table by child age group

This table reports:

- `n_records`: number of database rows  
- `% records`: relative frequency of rows  
- `population_sum`: total population represented  
- `% population`: percentage of the total 0-14 population  
- median population  
- population range 

```{r age-group-frequency}
age_group_summary <- pop_child %>%
  group_by(age_group) %>%
  summarise(
    n_records = n(),
    population_sum = sum(populacao, na.rm = TRUE),
    median_population = median(populacao, na.rm = TRUE),
    min_population = min(populacao, na.rm = TRUE),
    max_population = max(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    pct_records = n_records / sum(n_records) * 100,
    pct_population = population_sum / sum(population_sum) * 100,
    range_population = paste0(min_population, " - ", max_population)
  ) %>%
  select(
    age_group,
    n_records,
    pct_records,
    population_sum,
    pct_population,
    median_population,
    range_population
  ) %>%
  mutate(
    pct_records = round(pct_records, 2),
    pct_population = round(pct_population, 2),
    median_population = round(median_population, 1)
  )

kable(
  age_group_summary,
  caption = "Absolute and relative frequency by child age group"
)
```


# 5. Frequency table by year and child age group

```{r age-group-year-frequency}
age_group_year_summary <- pop_child %>%
  group_by(ano, age_group) %>%
  summarise(
    n_records = n(),
    population_sum = sum(populacao, na.rm = TRUE),
    median_population = median(populacao, na.rm = TRUE),
    min_population = min(populacao, na.rm = TRUE),
    max_population = max(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(ano) %>%
  mutate(
    pct_records_within_year = n_records / sum(n_records) * 100,
    pct_population_within_year = population_sum / sum(population_sum) * 100,
    range_population = paste0(min_population, " - ", max_population)
  ) %>%
  ungroup() %>%
  mutate(
    pct_records_within_year = round(pct_records_within_year, 2),
    pct_population_within_year = round(pct_population_within_year, 2),
    median_population = round(median_population, 1)
  )

kable(
  age_group_year_summary,
  caption = "Frequency by year and child age group"
)
```

# 6. Municipality-level child population percentage

Here we calculate the child population percentage for each municipality, year, and age group.


```{r municipality-child-percent}
total_pop_muni_year <- pop %>%
  group_by(cod_ibge, ano) %>%
  summarise(
    total_population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  )

child_pop_muni_year_agegroup <- pop_child %>%
  group_by(cod_ibge, ano, age_group) %>%
  summarise(
    child_population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    total_pop_muni_year,
    by = c("cod_ibge", "ano")
  ) %>%
  mutate(
    child_percent = child_population / total_population * 100
  )

glimpse(child_pop_muni_year_agegroup)
```

# 8. Quartiles and quintiles of child population percentage

Quartiles and quintiles are calculated within each year and age group.

```{r quartiles-quintiles}
child_percent_groups <- child_pop_muni_year_agegroup %>%
  group_by(ano, age_group) %>%
  mutate(
    child_percent_quartile = ntile(child_percent, 4),
    child_percent_quintile = ntile(child_percent, 5)
  ) %>%
  ungroup() %>%
  mutate(
    child_percent_quartile = factor(
      child_percent_quartile,
      levels = 1:4,
      labels = c("Q1 lowest", "Q2", "Q3", "Q4 highest")
    ),
    child_percent_quintile = factor(
      child_percent_quintile,
      levels = 1:5,
      labels = c("Q1 lowest", "Q2", "Q3", "Q4", "Q5 highest")
    )
  )

glimpse(child_percent_groups)
```


7. Quartile summary table

TThis reports:

- `n`: number of municipalities  
- `%`: percentage of municipalities in the quartile  
- median child population percentage  
- range of child population percentage  

```{r quartile-summary}
quartile_summary <- child_percent_groups %>%
  group_by(ano, age_group, child_percent_quartile) %>%
  summarise(
    n = n(),
    median_child_percent = median(child_percent, na.rm = TRUE),
    min_child_percent = min(child_percent, na.rm = TRUE),
    max_child_percent = max(child_percent, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(ano, age_group) %>%
  mutate(
    pct = n / sum(n) * 100,
    range_child_percent = paste0(
      round(min_child_percent, 2),
      "% - ",
      round(max_child_percent, 2),
      "%"
    )
  ) %>%
  ungroup() %>%
  mutate(
    pct = round(pct, 2),
    median_child_percent = round(median_child_percent, 2)
  ) %>%
  select(
    ano,
    age_group,
    child_percent_quartile,
    n,
    pct,
    median_child_percent,
    range_child_percent
  )

kable(
  quartile_summary,
  caption = "Quartile summary of municipality-level child population percentage"
)
```


# Plot-1. Line plot: child age-group percentage of total population over time

```{r line-child-percent-total, fig.width=9, fig.height=5}
child_percent_year_agegroup <- pop_child %>%
  group_by(ano, age_group) %>%
  summarise(
    child_population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    pop %>%
      group_by(ano) %>%
      summarise(
        total_population = sum(populacao, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "ano"
  ) %>%
  mutate(
    child_percent = child_population / total_population * 100
  )

child_percent_over_time_plot <- ggplot(
  child_percent_year_agegroup,
  aes(x = ano, y = child_percent, color = age_group, group = age_group)
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(
    title = "Child population percentage over time",
    subtitle = "Percentage of total population by child age group",
    x = "Year",
    y = "Percent of total population",
    color = "Age group"
  ) +
  theme_minimal(base_size = 13)

child_percent_over_time_plot

ggsave(
  filename = "/Users/abiramisomasundaram/Desktop/Brazil pesticides/Workspace/Plots/Population/child_population_percentage_over_time.jpeg",
  plot = child_percent_over_time_plot,
  width = 9,
  height = 5,
  dpi = 300
)
```


Map- Quartiles of child population percentage, ages 0-14

```{r}
sp_muni <- st_read(
  "/Users/abiramisomasundaram/Desktop/Brazil pesticides/Workspace/SHAPEFILES FOR MAPS-SAO PAULO BRAZIL/SP_Municipios_2025_with 6 digit/SP_Municipios_2025/SP_Municipios_2025.shp"
) %>%
  clean_names() %>%
  mutate(
    shapefile_code = as.character(cod_6),
    cod_ibge = str_pad(shapefile_code, width = 6, side = "left", pad = "0")
  )

names(sp_muni)

glimpse(sp_muni)


```

```{r}
# Select years to map
map_years <- c(2014, 2024)

# Total population by municipality and year
total_pop_muni_year_014 <- pop %>%
  group_by(cod_ibge, ano) %>%
  summarise(
    total_population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  )

# Child population age 0-14 by municipality and year
child_pop_014_muni_year <- pop_child %>%
  group_by(cod_ibge, ano) %>%
  summarise(
    child_population_0_14 = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    total_pop_muni_year_014,
    by = c("cod_ibge", "ano")
  ) %>%
  mutate(
    child_percent_0_14 = child_population_0_14 / total_population * 100
  )

# Calculate quartiles within each year
child_percent_014_quartiles <- child_pop_014_muni_year %>%
  filter(ano %in% map_years) %>%
  group_by(ano) %>%
  mutate(
    child_percent_014_quartile = ntile(child_percent_0_14, 4)
  ) %>%
  ungroup() %>%
  mutate(
    child_percent_014_quartile = factor(
      child_percent_014_quartile,
      levels = 1:4,
      labels = c("Q1 lowest", "Q2", "Q3", "Q4 highest")
    ),
    ano = factor(ano, levels = map_years)
  )

# Duplicate shapefile for the two years
sp_muni_years <- bind_rows(
  lapply(
    map_years,
    function(y) {
      sp_muni %>%
        mutate(
          ano = factor(y, levels = map_years)
        )
    }
  )
)

# Join quartile data to municipality geometry
quartile_map_014_sf <- sp_muni_years %>%
  left_join(
    child_percent_014_quartiles,
    by = c("cod_ibge", "ano")
  )

# Make the map
quartile_map_014_plot <- ggplot(quartile_map_014_sf) +
  geom_sf(
    aes(fill = child_percent_014_quartile),
    color = "white",
    linewidth = 0.1
  ) +
  facet_wrap(~ ano, ncol = 2) +
  scale_fill_manual(
    values = c(
      "Q1 lowest" = "#ffffcc",
      "Q2" = "#fd8d3c",
      "Q3" = "#f03b20",
      "Q4 highest" = "#bd0026"
    ),
    na.value = "grey85",
    drop = FALSE
  ) +
  labs(
    title = "Quartiles of child population percentage, ages 0-14",
    subtitle = "São Paulo municipalities, 2014 and 2024",
    fill = "Child population\npercentage quartile",
    caption = "Quartiles calculated within each year using municipality-level percentage aged 0-14."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

quartile_map_014_plot

ggsave(
  filename = "/Users/abiramisomasundaram/Desktop/Brazil pesticides/Workspace/Plots/Population/quartile_map_child_population_0_14_2014_2024.jpeg",
  plot = quartile_map_014_plot,
  width = 12,
  height = 6,
  dpi = 300
)

```



------------------------------------
Visualisations and Maps for Sex
------------------------------------
#Create sex-cleaned population object

```{r}
sex_values_check <- pop %>%
  count(sexo, sort = TRUE)

kable(
  sex_values_check,
  caption = "Original values of sex variable"
)

# Create standardized sex labels
pop_sex <- pop %>%
  mutate(
    sexo_lower = str_to_lower(str_trim(sexo)),
    sex_label = case_when(
      sexo_lower %in% c("f", "female", "feminino", "mulher", "2") |
        str_detect(sexo_lower, "fem") ~ "Female",
      
      sexo_lower %in% c("m", "male", "masculino", "homem", "1") |
        str_detect(sexo_lower, "masc") ~ "Male",
      
      TRUE ~ "Other/Unknown"
    ),
    sex_label = factor(
      sex_label,
      levels = c("Female", "Male", "Other/Unknown")
    )
  )

pop_child_sex <- pop_sex %>%
  filter(idade >= 0, idade <= 14)

pop_sex %>%
  count(sex_label, sort = TRUE) %>%
  kable(caption = "Standardized sex categories")



```

# Child population pyramid, ages 0–14, by sex

```{r}
map_years <- c(2014, 2024)

child_pyramid_percent_data <- pop_child_sex %>%
  filter(
    ano %in% map_years,
    sex_label %in% c("Female", "Male")
  ) %>%
  group_by(ano, idade, sex_label) %>%
  summarise(
    population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(ano) %>%
  mutate(
    total_child_population_year = sum(population, na.rm = TRUE),
    percent_child_population = population / total_child_population_year * 100
  ) %>%
  ungroup() %>%
  mutate(
    plot_percent = if_else(
      sex_label == "Male",
      -percent_child_population,
      percent_child_population
    ),
    ano = factor(ano, levels = map_years)
  )

child_pyramid_percent_plot <- ggplot(
  child_pyramid_percent_data,
  aes(
    x = factor(idade),
    y = plot_percent,
    fill = sex_label
  )
) +
  geom_col(width = 0.85) +
  coord_flip() +
  facet_wrap(~ ano, ncol = 2) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%")
  ) +
  scale_fill_manual(
    values = c(
      "Female" = "#E34A6F",
      "Male" = "#2C7FB8"
    )
  ) +
  labs(
    title = "Child population pyramid by sex, ages 0-14",
    subtitle = "Percentage of total child population within each year, São Paulo municipalities, 2014 and 2024",
    x = "Age",
    y = "Percent of child population aged 0-14",
    fill = "Sex",
    caption = "Male percentages are shown on the left; female percentages are shown on the right."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "bottom"
  )

child_pyramid_percent_plot

#save jpeg
ggsave(
  filename = "/Users/abiramisomasundaram/Desktop/Brazil pesticides/Workspace/Plots/Population/child_population_pyramid_percent_sex_2014_2024.jpeg",
  plot = child_pyramid_percent_plot,
  width = 11,
  height = 6,
  dpi = 300
)
```


```{r}
# Create municipality name if possible
name_candidates <- c(
  "nome_mun",
  "nm_mun",
  "nm_municip",
  "nm_municipio",
  "municipio",
  "name_muni",
  "nome"
)

muni_name_col <- intersect(name_candidates, names(sp_muni))[1]

sp_muni_named <- sp_muni %>%
  mutate(
    municipality_name = if (!is.na(muni_name_col)) {
      as.character(.data[[muni_name_col]])
    } else {
      cod_ibge
    }
  )

# Total municipal population by year
total_pop_muni_year_sexmap <- pop %>%
  group_by(cod_ibge, ano) %>%
  summarise(
    total_population = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  )

# Child population age 0-14 by municipality, year, and sex
sex_child_muni_year <- pop_child_sex %>%
  filter(sex_label %in% c("Female", "Male")) %>%
  group_by(cod_ibge, ano, sex_label) %>%
  summarise(
    sex_child_population_0_14 = sum(populacao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    total_pop_muni_year_sexmap,
    by = c("cod_ibge", "ano")
  ) %>%
  mutate(
    sex_child_percent_total = sex_child_population_0_14 / total_population * 100
  )

# Quartiles within each year and sex
sex_child_quartiles <- sex_child_muni_year %>%
  filter(ano %in% map_years) %>%
  group_by(ano, sex_label) %>%
  mutate(
    sex_child_percent_quartile = ntile(sex_child_percent_total, 4)
  ) %>%
  ungroup() %>%
  mutate(
    sex_child_percent_quartile = factor(
      sex_child_percent_quartile,
      levels = 1:4,
      labels = c("Q1 lowest", "Q2", "Q3", "Q4 highest")
    )
  )

# Duplicate shapefile by year and sex
sp_muni_years_sex <- bind_rows(
  lapply(
    map_years,
    function(y) {
      bind_rows(
        lapply(
          c("Female", "Male"),
          function(s) {
            sp_muni_named %>%
              mutate(
                ano = y,
                sex_label = s
              )
          }
        )
      )
    }
  )
)

# Join quartile data to geometry
sex_child_quartile_map_sf <- sp_muni_years_sex %>%
  left_join(
    sex_child_quartiles,
    by = c("cod_ibge", "ano", "sex_label")
  ) %>%
  mutate(
    ano = factor(ano, levels = map_years),
    sex_label = factor(sex_label, levels = c("Female", "Male"))
  )

# Map
sex_child_quartile_map_plot <- ggplot(sex_child_quartile_map_sf) +
  geom_sf(
    aes(fill = sex_child_percent_quartile),
    color = "white",
    linewidth = 0.1
  ) +
  facet_grid(sex_label ~ ano) +
  scale_fill_manual(
    values = c(
      "Q1 lowest" = "#ffffcc",
      "Q2" = "#fd8d3c",
      "Q3" = "#f03b20",
      "Q4 highest" = "#bd0026"
    ),
    na.value = "grey85",
    drop = FALSE
  ) +
  labs(
    title = "Quartiles of sex-specific child population percentage, ages 0-14",
    subtitle = "Percentage of total municipal population represented by female or male children, São Paulo, 2014 and 2024",
    fill = "Quartile",
    caption = "Quartiles calculated within each year and sex."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

sex_child_quartile_map_plot


``` 





```{r}
child_sex_ratio_muni_year <- sex_child_muni_year %>%
  filter(ano %in% map_years) %>%
  select(cod_ibge, ano, sex_label, sex_child_population_0_14) %>%
  pivot_wider(
    names_from = sex_label,
    values_from = sex_child_population_0_14,
    values_fill = 0
  ) %>%
  mutate(
    child_sex_ratio = if_else(
      Female > 0,
      Male / Female * 100,
      NA_real_
    )
  ) %>%
  group_by(ano) %>%
  mutate(
    child_sex_ratio_quartile = ntile(child_sex_ratio, 4)
  ) %>%
  ungroup() %>%
  mutate(
    child_sex_ratio_quartile = factor(
      child_sex_ratio_quartile,
      levels = 1:4,
      labels = c("Q1 lowest", "Q2", "Q3", "Q4 highest")
    )
  )

# Duplicate shapefile by year
sp_muni_years_ratio <- bind_rows(
  lapply(
    map_years,
    function(y) {
      sp_muni_named %>%
        mutate(
          ano = y
        )
    }
  )
)

# Join sex-ratio data to geometry
child_sex_ratio_map_sf <- sp_muni_years_ratio %>%
  left_join(
    child_sex_ratio_muni_year,
    by = c("cod_ibge", "ano")
  ) %>%
  mutate(
    ano = factor(ano, levels = map_years)
  )

# Map
child_sex_ratio_map_plot <- ggplot(child_sex_ratio_map_sf) +
  geom_sf(
    aes(fill = child_sex_ratio_quartile),
    color = "white",
    linewidth = 0.1
  ) +
  facet_wrap(~ ano, ncol = 2) +
  scale_fill_manual(
    values = c(
   "Q1 lowest" = "#EFF3FF",
    "Q2" = "#BDD7E7",
    "Q3" = "#6BAED6",
    "Q4 highest" = "#2171B5"
    ),
    na.value = "grey85",
    drop = FALSE
  ) +
  labs(
    title = "Quartiles of child sex ratio, ages 0-14",
    subtitle = "Boys per 100 girls by municipality, São Paulo, 2014 and 2024",
    fill = "Sex-ratio quartile",
    caption = "Quartiles calculated within each year. Sex ratio = male children / female children × 100."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

child_sex_ratio_map_plot

#save jpeg
ggsave(
  filename = "/Users/abiramisomasundaram/Desktop/Brazil pesticides/Workspace/Plots/Population/child_sex_ratio_quartile_map_2014_2024.jpeg",
  plot = child_sex_ratio_map_plot,
  width = 12,
  height = 6,
  dpi = 300
)


```



