---
title: "IVS"
output: html_document
date: "2026-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Packages

```{r}
library(readr)
library(arrow)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(DataExplorer)
library(knitr)
library(kableExtra)
```

##### Dataset

```{r}
ivs <- read_parquet("C:/Users/alyss/Documents/GitHub/intox_agro/resultados/contextual/ivs_municipios_sp_2010.parquet")
```

##### Cleaning

```{r}
#filter
ivs_clean <- ivs |>
  filter(label_sit_dom == "Total Situação de Domicílio")

ivs_clean <- ivs_clean |>
  filter(label_sexo == "Total Sexo")

ivs_clean <- ivs_clean |>
  filter(label_cor == "Total Cor")
```

```{r}
#variable class

class(ivs_clean$ivs_renda_e_trabalho)
class(ivs_clean$renda_per_capita)
class(ivs_clean$i_gini)
class(ivs_clean$t_analf_15m)
class(ivs_clean$ivs_capital_humano)
class(ivs_clean$t_sem_agua_esgoto)
class(ivs_clean$t_sem_lixo)
#make '-' NA

ivs_clean[ivs_clean == "-"] <- NA

#make numeric

ivs_clean <- ivs_clean %>%
  mutate(
    across(
      c(
        ivs_renda_e_trabalho,
        renda_per_capita,
        i_gini,
        t_analf_15m,
        ivs_capital_humano,
        t_sem_agua_esgoto,
        t_sem_lixo
      ),
      ~ as.numeric(gsub(",", ".", .))))
```

### Missingness

```{r}
plot_missing(ivs_clean)
```

# Income & Employment

## Income and Labor Sub-Index

### Distribution

```{r}

ivs_renda_e_trabalho_summary <- ivs_clean %>%
  select(ivs_renda_e_trabalho) %>%
  rename(value = ivs_renda_e_trabalho) %>%
  mutate(
    Quartile = ntile(desc(value), 4)
  ) %>%
  group_by(Quartile) %>%
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = "Income and Labor Sub-Index",
    Range = paste0(round(Min, 2), "-", round(Max, 2))
  ) %>%
  select(Variable, Quartile, n, pct, Median, Range)

#kable
ivs_renda_e_trabalho_summary %>%
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Income and Labor Sub-Index",
    col.names = c(
      "Income and Labor Sub-Index",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = ivs_renda_e_trabalho)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Per-Capita Income

### Distribution

```{r}

renda_per_capita_summary <- ivs_clean %>%
  select(renda_per_capita) %>%
  rename(value = renda_per_capita) %>%
  mutate(
    Quartile = ntile(desc(value), 4)
  ) %>%
  group_by(Quartile) %>%
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = "Per-Capita Income",
    Range = paste0(round(Min, 2), "-", round(Max, 2))
  ) %>%
  select(Variable, Quartile, n, pct, Median, Range)

#kable
renda_per_capita_summary %>%
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Per-Capita Income",
    col.names = c(
      "Per-Capita Income",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = renda_per_capita)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Income Inequality - Gini Coefficient

### Distribution

```{r}

i_gini_summary <- ivs_clean %>%
  select(i_gini) %>%
  rename(value = i_gini) %>%
  mutate(
    Quartile = ntile(desc(value), 4)
  ) %>%
  group_by(Quartile) %>%
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = "Gini Index",
    Range = paste0(round(Min, 2), "-", round(Max, 2))
  ) %>%
  select(Variable, Quartile, n, pct, Median, Range)

#kable
i_gini_summary %>%
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Gini Index",
    col.names = c(
      "Gini Index",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = i_gini)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Education

## Adult Literacy Rate = \> 15 years

### Distribution

```{r}

t_analf_15m_summary <- ivs_clean %>%
  select(t_analf_15m) %>%
  rename(value = t_analf_15m) %>%
  mutate(
    Quartile = ntile(desc(value), 4)
  ) %>%
  group_by(Quartile) %>%
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = "Adult Literacy Rate",
    Range = paste0(round(Min, 2), "-", round(Max, 2))
  ) %>%
  select(Variable, Quartile, n, pct, Median, Range)

#kable
t_analf_15m_summary %>%
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Adult Literacy Rate",
    col.names = c(
      "Adult Literacy Rate",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = t_analf_15m)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Human Capital Sub-Index

### Distribution

```{r}

ivs_capital_humano_summary <- ivs_clean %>%
  select(ivs_capital_humano) %>%
  rename(value = ivs_capital_humano) %>%
  mutate(
    Quartile = ntile(desc(value), 4)
  ) %>%
  group_by(Quartile) %>%
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = "Human Capital Sub-Index",
    Range = paste0(round(Min, 2), "-", round(Max, 2))
  ) %>%
  select(Variable, Quartile, n, pct, Median, Range)

#kable
ivs_capital_humano_summary %>%
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Human Capital Sub-Index",
    col.names = c(
      "Human Capital Sub-Index",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = ivs_capital_humano)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Sanitation & Environmental Exposure

## Households without Piped Water or Sewage

### Distribution

```{r}

t_sem_agua_esgoto_summary <- ivs_clean %>%
  select(t_sem_agua_esgoto) %>%
  rename(value = t_sem_agua_esgoto) %>%
  mutate(
    Quartile = ntile(desc(value), 4)
  ) %>%
  group_by(Quartile) %>%
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = "Households without Piped Water or Sewage",
    Range = paste0(round(Min, 2), "-", round(Max, 2))
  ) %>%
  select(Variable, Quartile, n, pct, Median, Range)

#kable
t_sem_agua_esgoto_summary %>%
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Households without Piped Water or Sewage",
    col.names = c(
      "Households without Piped Water or Sewage",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = t_sem_agua_esgoto)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Households without Garbage Collection

### Distribution

```{r}

t_sem_lixo_summary <- ivs_clean %>%
  select(t_sem_lixo) %>%
  rename(value = t_sem_lixo) %>%
  mutate(
    Quartile = ntile(desc(value), 4)
  ) %>%
  group_by(Quartile) %>%
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = "Households without Garbage Collection",
    Range = paste0(round(Min, 2), "-", round(Max, 2))
  ) %>%
  select(Variable, Quartile, n, pct, Median, Range)

#kable
t_sem_lixo_summary %>%
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Households without Garbage Collection",
    col.names = c(
      "Households without Garbage Collection",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = t_sem_lixo)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```
