---
title: "IVS"
output: html_document
date: "2026-06-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Packages

```{r}
library(readr)
library(arrow)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(DataExplorer)
library(knitr)
library(kableExtra)
```

##### Dataset

```{r}
ivs <- read_parquet("~/GitHub/intox_agro/resultados/contextual/ivs_municipios_sp_2010.parquet")
```

##### Cleaning

```{r}
#filter
ivs_clean <- ivs |>
  filter(label_sit_dom == "Total Situação de Domicílio")

ivs_clean <- ivs_clean |>
  filter(label_sexo == "Total Sexo")

ivs_clean <- ivs_clean |>
  filter(label_cor == "Total Cor")
```

```{r}
#variable class

class(ivs_clean$ivs_renda_e_trabalho)
class(ivs_clean$renda_per_capita)
class(ivs_clean$i_gini)
class(ivs_clean$t_analf_15m)
class(ivs_clean$ivs_capital_humano)
class(ivs_clean$t_sem_agua_esgoto)
class(ivs_clean$t_sem_lixo)
#make '-' NA

ivs_clean[ivs_clean == "-"] <- NA

#make numeric

ivs_clean <- ivs_clean |>
  mutate(
    across(
      c(
        ivs_renda_e_trabalho,
        renda_per_capita,
        i_gini,
        t_analf_15m,
        ivs_capital_humano,
        t_sem_agua_esgoto,
        t_sem_lixo
      ),
      ~ as.numeric(gsub(",", ".", .))))
```

### Missingness

```{r}
plot_missing(ivs_clean)
```

# Income & Employment

## Income and Labor Sub-Index

### Raw Frequency

```{r}

mean(ivs_clean$ivs_renda_e_trabalho)
median(ivs_clean$ivs_renda_e_trabalho)
```

#### Create Quartiles

```{r}
ivs_clean <- ivs_clean |>
  mutate(
    ivs_renda_e_trabalho_quartile = ntile(ivs_renda_e_trabalho, 4))
    
```

### Distribution

```{r}

ivs_renda_e_trabalho_summary <- ivs_clean |>
  group_by(ivs_renda_e_trabalho_quartile) |>
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(ivs_renda_e_trabalho, na.rm = TRUE),
    Min = min(ivs_renda_e_trabalho, na.rm = TRUE),
    Max = max(ivs_renda_e_trabalho, na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    Variable = "Income and Labor Sub-Index",
    Range = paste0(round(Min, 2), "-", round(Max, 2)) ) |>
  rename(Quartile = ivs_renda_e_trabalho_quartile) |>
  select(Variable, Quartile, n, pct, Median, Range)

#kable
ivs_renda_e_trabalho_summary |>
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Income and Labor Sub-Index",
    col.names = c(
      "Income and Labor Sub-Index",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) |>
  collapse_rows(columns = 1) |>
  kable_styling(full_width = FALSE)
```

### Plot

```{r}
ggplot(ivs_clean,
       aes(x = factor(ivs_renda_e_trabalho_quartile),
           y = ivs_renda_e_trabalho,
           fill = factor(ivs_renda_e_trabalho_quartile))) +
  geom_boxplot() +
  labs(
    title = "Income and Labor Sub-Index by Quartile",
    x = "Quartile",
    y = "Income and Labor Sub-Index",
    fill = "Quartile"
  ) +
  theme_minimal()
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = ivs_renda_e_trabalho)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Per-Capita Income

### Raw Frequencies

```{r}
mean(ivs_clean$renda_per_capita)
median(ivs_clean$renda_per_capita)
```

#### Create Quartiles

```{r}
ivs_clean <- ivs_clean |>
  mutate(
    renda_per_capita_quartile = ntile(renda_per_capita, 4))
```

### Distribution

```{r}

renda_per_capita_summary <- ivs_clean |>
  group_by(renda_per_capita_quartile) |>
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(renda_per_capita, na.rm = TRUE),
    Min = min(renda_per_capita, na.rm = TRUE),
    Max = max(renda_per_capita, na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    Variable = "Per-Capita Income",
    Range = paste0(round(Min, 2), "-", round(Max, 2))) |>
  rename(Quartile = renda_per_capita_quartile) |>
  select(Variable, Quartile, n, pct, Median, Range)

#kable
renda_per_capita_summary |>
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Per-Capita Income",
    col.names = c(
      "Per-Capita Income",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) |>
  collapse_rows(columns = 1) |>
  kable_styling(full_width = FALSE)
```

### Plot

```{r}
ggplot(ivs_clean,
       aes(x = factor(renda_per_capita_quartile),
           y = renda_per_capita,
           fill = factor(renda_per_capita_quartile))) +
  geom_boxplot() +
  labs(
    title = "Per-Capita Income by Quartile",
    x = "Quartile",
    y = "Per-Capita Income",
    fill = "Quartile"
  ) +
  theme_minimal()
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = renda_per_capita)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Income Inequality - Gini Coefficient

### Raw Frequencies

```{r}
mean(ivs_clean$i_gini)
median(ivs_clean$i_gini)
```

#### Create Quartiles

```{r}
ivs_clean <- ivs_clean |>
  mutate(
    i_gini_quartile = ntile(i_gini, 4))
```

### Distribution

```{r}

i_gini_summary <- ivs_clean |>
  group_by(i_gini_quartile) |>
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(i_gini, na.rm = TRUE),
    Min = min(i_gini, na.rm = TRUE),
    Max = max(i_gini, na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    Variable = "Gini Index",
    Range = paste0(round(Min, 2), "-", round(Max, 2))) |>
  rename(Quartile = i_gini_quartile) |>
  select(Variable, Quartile, n, pct, Median, Range)

#kable
i_gini_summary |>
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Gini Index",
    col.names = c(
      "Gini Index",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) |>
  collapse_rows(columns = 1) |>
  kable_styling(full_width = FALSE)
```

### Plot

```{r}
ggplot(ivs_clean,
       aes(x = factor(i_gini_quartile),
           y = i_gini,
           fill = factor(i_gini_quartile))) +
  geom_boxplot() +
  labs(
    title = "Gini Index by Quartile",
    x = "Quartile",
    y = "Gini Index",
    fill = "Quartile"
  ) +
  theme_minimal()
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = i_gini)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Education

## Adult Iliteracy Rate = \> 15 years

### Raw Frequencies

```{r}
mean(ivs_clean$t_analf_15m)
median(ivs_clean$t_analf_15m)
```

#### Create Quartiles

```{r}
ivs_clean <- ivs_clean |>
  mutate(
    t_analf_15m_quartile = ntile(t_analf_15m, 4))
```

### Distribution

```{r}

t_analf_15m_summary <- ivs_clean |>
  group_by(t_analf_15m_quartile) |>
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(t_analf_15m, na.rm = TRUE),
    Min = min(t_analf_15m, na.rm = TRUE),
    Max = max(t_analf_15m, na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    Variable = "Adult Iliteracy Rate",
    Range = paste0(round(Min, 2), "-", round(Max, 2))) |>
  rename(Quartile = t_analf_15m_quartile) |>
  select(Variable, Quartile, n, pct, Median, Range)

#kable
t_analf_15m_summary |>
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Adult Iliteracy Rate",
    col.names = c(
      "Adult Iliteracy Ratee",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) |>
  collapse_rows(columns = 1) |>
  kable_styling(full_width = FALSE)
```

### Plot

```{r}
ggplot(ivs_clean,
       aes(x = factor(t_analf_15m_quartile),
           y = t_analf_15m,
           fill = factor(t_analf_15m_quartile))) +
  geom_boxplot() +
  labs(
    title = "Adult Iliteracy Rate by Quartile",
    x = "Quartile",
    y = "Adult Iliteracy Rate",
    fill = "Quartile"
  ) +
  theme_minimal()
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = t_analf_15m)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Human Capital Sub-Index

### Raw Frequencies

```{r}
mean(ivs_clean$ivs_capital_humano)
median(ivs_clean$ivs_capital_humano)
```

#### Create Quartiles

```{r}
ivs_clean <- ivs_clean |>
  mutate(
    ivs_capital_humano_quartile = ntile(ivs_capital_humano, 4))
```

### Distribution

```{r}

ivs_capital_humano_summary <- ivs_clean |>
  group_by(ivs_capital_humano_quartile) |>
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(ivs_capital_humano, na.rm = TRUE),
    Min = min(ivs_capital_humano, na.rm = TRUE),
    Max = max(ivs_capital_humano, na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    Variable = "Human Capital Sub-Index",
    Range = paste0(round(Min, 2), "-", round(Max, 2))) |>
  rename(Quartile = ivs_capital_humano_quartile) |>
  select(Variable, Quartile, n, pct, Median, Range)

#kable
ivs_capital_humano_summary |>
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Human Capital Sub-Index",
    col.names = c(
      "Human Capital Sub-Index",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) |>
  collapse_rows(columns = 1) |>
  kable_styling(full_width = FALSE)
```

### Plot

```{r}
ggplot(ivs_clean,
       aes(x = factor(ivs_capital_humano_quartile),
           y = ivs_capital_humano,
           fill = factor(ivs_capital_humano_quartile))) +
  geom_boxplot() +
  labs(
    title = "Human Capital Sub-Index By Quartile",
    x = "Quartile",
    y = "Human Capital Sub-Index",
    fill = "Quartile"
  ) +
  theme_minimal()
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = ivs_capital_humano)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Sanitation & Environmental Exposure

## Households without Piped Water or Sewage

### Raw Frequencies

```{r}
mean(ivs_clean$t_sem_agua_esgoto)
median(ivs_clean$t_sem_agua_esgoto)
```

#### Create Quartiles

```{r}
ivs_clean <- ivs_clean |>
  mutate(
    t_sem_agua_esgoto_quartile = ntile(t_sem_agua_esgoto, 4))
```

### Distribution

```{r}

t_sem_agua_esgoto_summary <- ivs_clean |>
  group_by(t_sem_agua_esgoto_quartile) |>
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(t_sem_agua_esgoto, na.rm = TRUE),
    Min = min(t_sem_agua_esgoto, na.rm = TRUE),
    Max = max(t_sem_agua_esgoto, na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    Variable = "Households without Piped Water or Sewage",
    Range = paste0(round(Min, 2), "-", round(Max, 2))) |>
  rename(Quartile = t_sem_agua_esgoto_quartile) |>
  select(Variable, Quartile, n, pct, Median, Range)

#kable
t_sem_agua_esgoto_summary |>
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Households without Piped Water or Sewage",
    col.names = c(
      "Households without Piped Water or Sewage",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) |>
  collapse_rows(columns = 1) |>
  kable_styling(full_width = FALSE)
```

### Plot

```{r}
ggplot(ivs_clean,
       aes(x = factor(t_sem_agua_esgoto_quartile),
           y = t_sem_agua_esgoto,
           fill = factor(t_sem_agua_esgoto_quartile))) +
  geom_boxplot() +
  labs(
    title = "Households without Piped Water or Sewage By Quartile",
    x = "Quartile",
    y = "Households without Piped Water or Sewage",
    fill = "Quartile"
  ) +
  theme_minimal()
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = t_sem_agua_esgoto)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Households without Garbage Collection

### Raw Frequencies

```{r}
mean(ivs_clean$t_sem_lixo)
median(ivs_clean$t_sem_lixo)
```

#### Create Quartiles

```{r}
ivs_clean <- ivs_clean |>
  mutate(
    t_sem_lixo_quartile = ntile(t_sem_lixo, 4))
```

### Distribution

```{r}

t_sem_lixo_summary <- ivs_clean |>
  group_by(t_sem_lixo_quartile) |>
  summarise(
    n = n(),
    pct = round(100 * n() / nrow(ivs_clean), 1),
    Median = median(t_sem_lixo, na.rm = TRUE),
    Min = min(t_sem_lixo, na.rm = TRUE),
    Max = max(t_sem_lixo, na.rm = TRUE),
    .groups = "drop") |>
  mutate(
    Variable = "Households without Garbage Collectio",
    Range = paste0(round(Min, 2), "-", round(Max, 2))) |>
  rename(Quartile = t_sem_lixo_quartile) |>
  select(Variable, Quartile, n, pct, Median, Range)

#kable
t_sem_lixo_summary |>
  kable(
    digits = 1,
    caption = "Distribution of Municipalities by Households without Garbage Collection",
    col.names = c(
      "Households without Garbage Collection",
      "Quartile",
      "n",
      "%",
      "Median (%)",
      "Range"
    )
  ) |>
  collapse_rows(columns = 1) |>
  kable_styling(full_width = FALSE)
```

### Plot

```{r}
ggplot(ivs_clean,
       aes(x = factor(t_sem_lixo_quartile),
           y = t_sem_lixo,
           fill = factor(t_sem_lixo_quartile))) +
  geom_boxplot() +
  labs(
    title = "Households without Garbage Collection By Quartile",
    x = "Quartile",
    y = "Households without Garbage Collection",
    fill = "Quartile"
  ) +
  theme_minimal()
```

### Normality

```{r}
ggplot(ivs_clean, aes(x = t_sem_lixo)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```
