---
title: "Base - Agricultural Analysis"
output: html_document
date: "2026-06-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Packages

```{r}
library(readr)
library(arrow)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(DataExplorer)
library(knitr)
library(kableExtra)
```

##### Dataset

```{r}
consolidada <- read_parquet("~/GitHub/intox_agro/resultados/base_consolidada_sp_2014_2024.parquet")
```

## Missingness

```{r}
plot_missing(
  consolidada[, c(
    "sisagua_n_pesticidas_detect",
    "sisagua_pct_deteccao",
    "pam_area_colhida_ha",
    "pam_valor_prod_mil_reais")])
```

```{r}
#year missingness
consolidada %>%
  group_by(ano) %>%
  summarise(
    pct_missing_detect =
      mean(is.na(sisagua_pct_deteccao)) * 100,
    pct_missing_area =
      mean(is.na(pam_area_colhida_ha)) * 100,
    pct_missing_value =
      mean(is.na(pam_valor_prod_mil_reais)) * 100
  )

#municipality missingness
consolidada %>%
  filter(is.na(sisagua_pct_deteccao)) %>%
  distinct(cod_ibge)

consolidada %>%
  filter(is.na(pam_area_colhida_ha)) %>%
  summarise(n_municipalities = n_distinct(cod_ibge))

#trends in missingness
consolidada %>%
  group_by(is.na(sisagua_pct_deteccao)) %>%
  summarise(
    median_population = median(pop_total, na.rm = TRUE),
    median_income = median(renda_per_capita, na.rm = TRUE),
    n = n())

consolidada %>%
  group_by(is.na(pam_valor_prod_mil_reais)) %>%
  summarise(
    median_population = median(pop_total, na.rm = TRUE),
    median_income = median(renda_per_capita, na.rm = TRUE),
    n = n())


#identical missingness
all(
  is.na(consolidada$pam_area_colhida_ha) ==
    is.na(consolidada$pam_valor_prod_mil_reais))
```

`sisagua_pct_deteccao` is missing \>10% in years 2014-2017 and 2023-2024, in a total of 484 municipalities. `pam_area_colhida_ha` and `pam_valor_prod_mil_reais` have consistent missingness around 6-7% - and are missing in the exact same municipalities. The median population of municipalities with missing data for `sisagua_pct_deteccao` is 8277 compared to the median population of non-missing municipalities being 15975. Income does not appear to be a trend (695.67 in non-missing vs 636 in missing). On the other hand, PAM missingness is higher in municipalities with larger median populations (159572 in missing vs 11668 in non-missing). Additionally, missingness is concentrated in municipalities with higher median income (767.85 in missing vs 680.56 in non-missing).

# Samples with at least one quantifiable pesticide detection

## Raw Frequencies

```{r}

#table
detect_raw_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    mean_amostras = mean(sisagua_pct_deteccao, na.rm = TRUE),
    median_amostras = median(sisagua_pct_deteccao, na.rm = TRUE),
    .groups = "drop")

detect_raw_summary

#plot
detect_raw_plot <- consolidada %>%
  group_by(ano) %>%
  summarise(
    Mean = mean(sisagua_pct_deteccao, na.rm = TRUE),
    Median = median(sisagua_pct_deteccao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "Value")

ggplot(detect_raw_plot,
       aes(x = ano, y = Value, color = Statistic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Mean and Median of % Samples with Quantifiable Detection by Year",
    x = "Year",
    y = "% of Detected Samples",
    color = NULL
  ) +
  theme_minimal()
```

### Create Quartiles

```{r}
consolidada <- consolidada %>%
  group_by(ano) %>%
  mutate(
    sisagua_pct_deteccao_quartile =
      ntile(sisagua_pct_deteccao, 4)
  ) %>%
  ungroup()
```

## Table

```{r}

detect_summary <- consolidada %>%
  select(
    ano,
    sisagua_pct_deteccao,
    sisagua_pct_deteccao_quartile
  ) %>%
  rename(
    value = sisagua_pct_deteccao,
    Quartile = sisagua_pct_deteccao_quartile
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

detect_summary

#kable

detect_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Quartile",
      "n",
      "Median",
      "Range"
    ),
    caption = "% Detected Samples in Quartiles by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}

ggplot(detect_summary,
       aes(x = ano,
           y = Median,
           color = factor(Quartile),
           group = Quartile)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Median % Samples with Quantifiable Detection in Quartiles by Year",
    x = "Year",
    y = "Median % Samples with Quantifiable Detection",
    color = "Quartile"
  ) +
  theme_minimal()

```

Here we can see that a majority of municipalities do not detect quantifiable levels of pesticides in their samples. However, among samples in the top 25% and 50% of samples with quantifiable levels of pesticides, we see that 100% of samples tested from 2014-2016 had quantifiable levels of pesticides, dropping to below 25% from 2019-2022 and rising back up to above 50% 2023-2024.

## Normality

```{r}
ggplot(consolidada, aes(x = sisagua_pct_deteccao)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Number of distinct pesticide compounds detected in quantifiable amounts

## Raw Frequencies

```{r}
#table
distinct_raw_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    mean_amostras = mean(sisagua_n_pesticidas_detect, na.rm = TRUE),
    median_amostras = median(sisagua_n_pesticidas_detect, na.rm = TRUE),
    .groups = "drop")

distinct_raw_summary

#plot
distinct_plot <- consolidada %>%
  group_by(ano) %>%
  summarise(
    Mean = mean(sisagua_n_pesticidas_detect, na.rm = TRUE),
    Median = median(sisagua_n_pesticidas_detect, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "Value")

ggplot(distinct_plot,
       aes(x = ano, y = Value, color = Statistic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Mean and Median of # Distinct Pesticides Detected by Year",
    x = "Year",
    y = "# Distinct Pesticides Detected",
    color = NULL
  ) +
  theme_minimal()
```

### Create Quartiles

```{r}
consolidada <- consolidada %>%
  group_by(ano) %>%
  mutate(
    sisagua_n_pesticidas_detect_quartile =
      ntile(sisagua_n_pesticidas_detect, 4)
  ) %>%
  ungroup()
```

## Table

```{r}
distinct_summary <- consolidada %>%
  select(
    ano,
    sisagua_n_pesticidas_detect,
    sisagua_n_pesticidas_detect_quartile
  ) %>%
  rename(
    value = sisagua_n_pesticidas_detect,
    Quartile = sisagua_n_pesticidas_detect_quartile
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

distinct_summary

#kable

distinct_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Quartile",
      "n",
      "Median",
      "Range"
    ),
    caption = "# Distinct Pesticides Detected in Quartiles by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}
ggplot(distinct_summary,
       aes(x = ano,
           y = Median,
           color = factor(Quartile),
           group = Quartile)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Median # Distinct Pesticides Detected by Quartile and Year",
    x = "Year",
    y = "Median % Detected Samples",
    color = "Quartile"
  ) +
  theme_minimal()
```

From 2014-2022, 27 distinct pesticides were detected in the top 25% of municipalities.

## Normality

```{r}
ggplot(consolidada, aes(x = sisagua_n_pesticidas_detect)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Total harvested area across all crops (hectares)

## Raw Frequencies

```{r}
#table
area_raw_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    mean_amostras = mean(pam_area_colhida_ha, na.rm = TRUE),
    median_amostras = median(pam_area_colhida_ha, na.rm = TRUE),
    .groups = "drop")

area_raw_summary

#plot
area_raw_plot <- consolidada %>%
  group_by(ano) %>%
  summarise(
    Mean = mean(pam_area_colhida_ha, na.rm = TRUE),
    Median = median(pam_area_colhida_ha, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "Value")

ggplot(distinct_plot,
       aes(x = ano, y = Value, color = Statistic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Mean and Median of Harvested Area by Year",
    x = "Year",
    y = "Are Harvested (hectares)",
    color = NULL
  ) +
  theme_minimal()
```

### Create Quartiles

```{r}
consolidada <- consolidada %>%
group_by(ano) %>%
  mutate(
    pam_area_colhida_ha_quartile =
      ntile(pam_area_colhida_ha, 4)
  ) %>%
  ungroup()
```

## Table

```{r}
area_summary <- consolidada %>%
  select(
    ano,
    pam_area_colhida_ha,
    pam_area_colhida_ha_quartile
  ) %>%
  rename(
    value = pam_area_colhida_ha,
    Quartile = pam_area_colhida_ha_quartile
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

area_summary

#kable

area_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Quartile",
      "n",
      "Median",
      "Range"
    ),
    caption = "Harvested Area in Quartiles by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}
ggplot(area_summary,
       aes(x = ano,
           y = Median,
           color = factor(Quartile),
           group = Quartile)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
    breaks = seq(0, 40000, by = 5000)
  ) +
  labs(
    title = "Harvested Area in Quartiles by Year",
    x = "Year",
    y = "Harvested Area (hectares)",
    color = "Quartile"
  ) +
  theme_minimal()
```

The median area of harvested land in the top 25% municipalities increases from 30,000 in 2014 hectares to 34,218 hectares in 2023 and drops to 33,518 in 2024. The second-highest 25% of municipalities see steady median areas of harvested land between 2014-2024, ranging from 11,963.5 hectares to 13,445.5 hectares.

## Normality

```{r}
ggplot(consolidada, aes(x = pam_area_colhida_ha)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Total agricultural production value, all crops (thousands BRL, current prices)

## Raw Frequencies

```{r}
#table
valor_raw_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    mean_amostras = mean(pam_valor_prod_mil_reais, na.rm = TRUE),
    median_amostras = median(pam_valor_prod_mil_reais, na.rm = TRUE),
    .groups = "drop")

area_raw_summary

#plot
valor_raw_plot <- consolidada %>%
  group_by(ano) %>%
  summarise(
    Mean = mean(pam_valor_prod_mil_reais, na.rm = TRUE),
    Median = median(pam_valor_prod_mil_reais, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "Value")

ggplot(valor_raw_plot,
       aes(x = ano, y = Value, color = Statistic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Mean and Median Of Total Agricultural Production Value by Year",
    x = "Year",
    y = "Total Agricultural Production Value",
    color = NULL
  ) +
  theme_minimal()
```

### Create Quartiles

```{r}
consolidada <- consolidada %>%
  group_by(ano) %>%
  mutate(
    pam_valor_prod_mil_reais_quartile =
      ntile(pam_valor_prod_mil_reais, 4)
  ) %>%
  ungroup()
```

## Table

```{r}
valor_summary <- consolidada %>%
  select(
    ano,
    pam_valor_prod_mil_reais,
    pam_valor_prod_mil_reais_quartile
  ) %>%
  rename(
    value = pam_valor_prod_mil_reais,
    Quartile = pam_valor_prod_mil_reais_quartile
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

valor_summary

#kable

valor_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Quartile",
      "n",
      "Median",
      "Range"
    ),
    caption = "Total Agricultural Production Value in Quartiles by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}
ggplot(valor_summary,
       aes(x = ano,
           y = Median,
           color = factor(Quartile),
           group = Quartile)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Median Total Agricultural Production Value in Quartiles by Year",
    x = "Year",
    y = "Median Total Agricultural Production Value (BRL)",
    color = "Quartile"
  ) +
  theme_minimal()
```

The median total agricultural production value increases between 2014 and 2024 for all municipalities. The top 25% see this increase from 132,040.5 BRL in 2014 to 416,699.0 BRL in 2024. s

## Normality

```{r}
ggplot(consolidada, aes(x = pam_valor_prod_mil_reais)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

#### Save Dataset

```{r}
agricultural_gis <- consolidada %>%
  select(
    cod_ibge,
    sisagua_pct_deteccao_quartile,
    sisagua_n_pesticidas_detect_quartile,
    pam_area_colhida_ha_quartile,
    pam_valor_prod_mil_reais_quartile
  )


#save separate dataset per year

for (yr in 2014:2024) {
  consolidada %>%
    filter(ano == yr) %>%
    write_csv(paste0("agricultural_", yr, ".csv"))
}

```
