---
title: "Base - Agricultural Analysis"
output: html_document
date: "2026-06-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Packages

```{r}
library(readr)
library(arrow)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(DataExplorer)
library(knitr)
library(kableExtra)
library(classInt)
```

##### Dataset

```{r}
consolidada <- read_parquet("~/GitHub/intox_agro/resultados/base_consolidada_sp_2014_2024.parquet")
```

## Missingness

```{r}
plot_missing(
  consolidada[, c(
    "sisagua_n_pesticidas_detect",
    "sisagua_pct_deteccao",
    "pam_area_colhida_ha",
    "pam_valor_prod_mil_reais")])
```

```{r}
#year missingness
missing_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    `Detection (%)` = mean(is.na(sisagua_pct_deteccao)) * 100,
    `Harvested Area (%)` = mean(is.na(pam_area_colhida_ha)) * 100,
    `Crop Value (%)` = mean(is.na(pam_valor_prod_mil_reais)) * 100,
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = -ano,
    names_to = "Variable",
    values_to = "Percent_Missing"
  )

ggplot(missing_summary,
       aes(x = ano, y = Percent_Missing,
           color = Variable, group = Variable)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = sort(unique(missing_summary$ano))) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, 10)
  ) +
  labs(
    title = "Percentage of Missing Values by Year",
    x = "Year",
    y = "Missing (%)",
    color = "Variable"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

#municipality missingness
consolidada %>%
  filter(is.na(sisagua_pct_deteccao)) %>%
  distinct(cod_ibge)

consolidada %>%
  filter(is.na(pam_area_colhida_ha)) %>%
  summarise(n_municipalities = n_distinct(cod_ibge))

#trends in missingness
consolidada %>%
  group_by(is.na(sisagua_pct_deteccao)) %>%
  summarise(
    median_population = median(pop_total, na.rm = TRUE),
    median_income = median(renda_per_capita, na.rm = TRUE),
    n = n())

consolidada %>%
  group_by(is.na(pam_valor_prod_mil_reais)) %>%
  summarise(
    median_population = median(pop_total, na.rm = TRUE),
    median_income = median(renda_per_capita, na.rm = TRUE),
    n = n())


#identical missingness
all(
  is.na(consolidada$pam_area_colhida_ha) ==
    is.na(consolidada$pam_valor_prod_mil_reais))
```

`sisagua_pct_deteccao` is missing \>10% in years 2014-2017 and 2023-2024, in a total of 484 municipalities. `pam_area_colhida_ha` and `pam_valor_prod_mil_reais` have consistent missingness around 6-7% - and are missing in the exact same municipalities. The median population of municipalities with missing data for `sisagua_pct_deteccao` is 8277 compared to the median population of non-missing municipalities being 15975. Income does not appear to be a trend (695.67 in non-missing vs 636 in missing). On the other hand, PAM missingness is higher in municipalities with larger median populations (159572 in missing vs 11668 in non-missing). Additionally, missingness is concentrated in municipalities with higher median income (767.85 in missing vs 680.56 in non-missing).

# Samples with at least one quantifiable pesticide detection

## Raw Frequencies

```{r}

#table
detect_raw_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    mean_amostras = mean(sisagua_pct_deteccao, na.rm = TRUE),
    median_amostras = median(sisagua_pct_deteccao, na.rm = TRUE),
    .groups = "drop")

detect_raw_summary

#plot
detect_raw_plot <- consolidada %>%
  group_by(ano) %>%
  summarise(
    Mean = mean(sisagua_pct_deteccao, na.rm = TRUE),
    Median = median(sisagua_pct_deteccao, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "Value")

ggplot(detect_raw_plot,
       aes(x = ano, y = Value, color = Statistic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Mean and Median of % Samples with Quantifiable Detection by Year",
    x = "Year",
    y = "% of Detected Samples",
    color = NULL
  ) +
  theme_minimal()
```

### Create Quartiles

```{r}

#breaks_deteccao <- classIntervals(
 # consolidada$sisagua_pct_deteccao,
  #n = 4,
  #style = "equal")$brks

#consolidada <- consolidada %>%
 # mutate(
  #  sisagua_pct_deteccao_quartile = cut(
   #   sisagua_pct_deteccao,
    #  breaks = c(0, 25, 50, 75, 100),
     # include.lowest = TRUE,
      #labels = 1:4))
quantile(consolidada$sisagua_pct_deteccao,
         probs = c(0, 0.25, 0.5, 0.75, 1),
         na.rm = TRUE)


consolidada <- consolidada %>%
  mutate(
    sisagua_pct_deteccao_q = ntile(sisagua_pct_deteccao, 4)
  )

```

## Table

```{r}

detect_summary <- consolidada %>%
  select(
    ano,
    sisagua_pct_deteccao,
    sisagua_pct_deteccao_q
  ) %>%
  rename(
    value = sisagua_pct_deteccao,
    Quartile = sisagua_pct_deteccao_q
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

detect_summary

#kable

detect_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Category",
      "n",
      "Median",
      "Range"
    ),
    caption = "% Detected Samples by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}

ggplot(detect_summary,
       aes(x = ano,
           y = Median,
           color = factor(Quartile),
           group = Quartile)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Median % Samples with Quantifiable Detection by Year",
    x = "Year",
    y = "Median % Samples with Quantifiable Detection",
    color = "Category"
  ) +
  theme_minimal()

#Distribution
category_counts <- consolidada %>%
  group_by(ano, sisagua_pct_deteccao_quartile) %>%
  summarise(
    n_municipalities = n(),
    .groups = "drop"
  )

category_counts %>%
  ggplot(aes(x = ano,
             y = n_municipalities,
             color = sisagua_pct_deteccao_quartile,
             group = sisagua_pct_deteccao_quartile)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(category_counts$ano)) +
  labs(
    x = "Year",
    y = "Number of municipalities",
    color = "Detection category"
  ) +
  theme_minimal()
```

Median quantifiable pesticide detection percentages remained relatively stable between 2014 and 2024. Municipalities in the highest detection category consistently had median values near 100%, while those in the lowest category were near 0% from 2014-2024. We also see in the graph showing the number of municipalities in each category over time, that there are more municipalities joining the lowest levels of detections (category 1) compared to the other categories, while NA also increases.

## Normality

```{r}
ggplot(consolidada, aes(x = sisagua_pct_deteccao)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Number of distinct pesticide compounds detected in quantifiable amounts

## Raw Frequencies

```{r}
#table
distinct_raw_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    mean_amostras = mean(sisagua_n_pesticidas_detect, na.rm = TRUE),
    median_amostras = median(sisagua_n_pesticidas_detect, na.rm = TRUE),
    .groups = "drop")

distinct_raw_summary

#plot
distinct_plot <- consolidada %>%
  group_by(ano) %>%
  summarise(
    Mean = mean(sisagua_n_pesticidas_detect, na.rm = TRUE),
    Median = median(sisagua_n_pesticidas_detect, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "Value")

ggplot(distinct_plot,
       aes(x = ano, y = Value, color = Statistic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Mean and Median of # Distinct Pesticides Detected by Year",
    x = "Year",
    y = "# Distinct Pesticides Detected",
    color = NULL
  ) +
  theme_minimal()
```

### Create Quartiles

```{r}
#consolidada <- consolidada %>%
 # group_by(ano) %>%
 # mutate(
 #   sisagua_n_pesticidas_detect_quartile =
 #     ntile(sisagua_n_pesticidas_detect, 4)
 # ) %>%
#  ungroup()

breaks_detect <- classIntervals(
  consolidada$sisagua_n_pesticidas_detect,
  n = 4,
  style = "equal")$brks

consolidada <- consolidada %>%
  mutate(
    sisagua_n_pesticidas_detect_q = cut(
      sisagua_n_pesticidas_detect,
      breaks = c(0, 25, 50, 75, 100),
      include.lowest = TRUE,
      labels = 1:4))
```

## Table

```{r}
distinct_summary <- consolidada %>%
  select(
    ano,
    sisagua_n_pesticidas_detect,
    sisagua_n_pesticidas_detect_q
  ) %>%
  rename(
    value = sisagua_n_pesticidas_detect,
    Quartile = sisagua_n_pesticidas_detect_q
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

distinct_summary

#kable

distinct_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Category",
      "n",
      "Median",
      "Range"
    ),
    caption = "# Distinct Pesticides Detected by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}
ggplot(distinct_summary,
       aes(x = ano,
           y = Median,
           color = factor(Quartile),
           group = Quartile)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Median # Distinct Pesticides Detected by Quartile and Year",
    x = "Year",
    y = "Median % Detected Samples",
    color = "Quartile"
  ) +
  theme_minimal()
```

From 2014-2022, 27 distinct pesticides were detected in the top 25% of municipalities.

## Normality

```{r}
ggplot(consolidada, aes(x = sisagua_n_pesticidas_detect)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Total harvested area across all crops (hectares)

## Raw Frequencies

```{r}
#table
area_raw_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    mean_amostras = mean(pam_area_colhida_ha, na.rm = TRUE),
    median_amostras = median(pam_area_colhida_ha, na.rm = TRUE),
    .groups = "drop")

area_raw_summary

#plot
area_raw_plot <- consolidada %>%
  group_by(ano) %>%
  summarise(
    Mean = mean(pam_area_colhida_ha, na.rm = TRUE),
    Median = median(pam_area_colhida_ha, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "Value")

ggplot(distinct_plot,
       aes(x = ano, y = Value, color = Statistic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Mean and Median of Harvested Area by Year",
    x = "Year",
    y = "Are Harvested (hectares)",
    color = NULL
  ) +
  theme_minimal()
```

### Create Quartiles

```{r}
#consolidada <- consolidada %>%
#group_by(ano) %>%
 # mutate(
  #  pam_area_colhida_ha_quartile =
  #    ntile(pam_area_colhida_ha, 4)
  #) %>%
 # ungroup()
```

## Table

```{r}
area_summary <- consolidada %>%
  select(
    ano,
    pam_area_colhida_ha,
    pam_area_colhida_ha_quartile
  ) %>%
  rename(
    value = pam_area_colhida_ha,
    Quartile = pam_area_colhida_ha_quartile
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

area_summary

#kable

area_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Quartile",
      "n",
      "Median",
      "Range"
    ),
    caption = "Harvested Area in Quartiles by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}
ggplot(area_summary,
       aes(x = ano,
           y = Median,
           color = factor(Quartile),
           group = Quartile)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(
    breaks = seq(0, 40000, by = 5000)
  ) +
  labs(
    title = "Harvested Area in Quartiles by Year",
    x = "Year",
    y = "Harvested Area (hectares)",
    color = "Quartile"
  ) +
  theme_minimal()
```

The median area of harvested land in the top 25% municipalities increases from 30,000 in 2014 hectares to 34,218 hectares in 2023 and drops to 33,518 in 2024. The second-highest 25% of municipalities see steady median areas of harvested land between 2014-2024, ranging from 11,963.5 hectares to 13,445.5 hectares.

## Normality

```{r}
ggplot(consolidada, aes(x = pam_area_colhida_ha)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Total agricultural production value, all crops (thousands BRL, current prices)

## Raw Frequencies

```{r}
#table
valor_raw_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    mean_amostras = mean(pam_valor_prod_mil_reais, na.rm = TRUE),
    median_amostras = median(pam_valor_prod_mil_reais, na.rm = TRUE),
    .groups = "drop")

area_raw_summary

#plot
valor_raw_plot <- consolidada %>%
  group_by(ano) %>%
  summarise(
    Mean = mean(pam_valor_prod_mil_reais, na.rm = TRUE),
    Median = median(pam_valor_prod_mil_reais, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Mean, Median),
    names_to = "Statistic",
    values_to = "Value")

ggplot(valor_raw_plot,
       aes(x = ano, y = Value, color = Statistic)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Mean and Median Of Total Agricultural Production Value by Year",
    x = "Year",
    y = "Total Agricultural Production Value",
    color = NULL
  ) +
  theme_minimal()
```

### Create Quartiles

```{r}
#consolidada <- consolidada %>%
  #group_by(ano) %>%
 # mutate(
  #  pam_valor_prod_mil_reais_quartile =
    #  ntile(pam_valor_prod_mil_reais, 4)
 # ) %>%
  #ungroup()


```

## Table

```{r}
valor_summary <- consolidada %>%
  select(
    ano,
    pam_valor_prod_mil_reais,
    pam_valor_prod_mil_reais_quartile
  ) %>%
  rename(
    value = pam_valor_prod_mil_reais,
    Quartile = pam_valor_prod_mil_reais_quartile
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

valor_summary

#kable

valor_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Quartile",
      "n",
      "Median",
      "Range"
    ),
    caption = "Total Agricultural Production Value in Quartiles by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}
ggplot(valor_summary,
       aes(x = ano,
           y = Median,
           color = factor(Quartile),
           group = Quartile)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Median Total Agricultural Production Value in Quartiles by Year",
    x = "Year",
    y = "Median Total Agricultural Production Value (BRL)",
    color = "Quartile"
  ) +
  theme_minimal()
```

The median total agricultural production value increases between 2014 and 2024 for all municipalities. The top 25% see this increase from 132,040.5 BRL in 2014 to 416,699.0 BRL in 2024. s

## Normality

```{r}
ggplot(consolidada, aes(x = pam_valor_prod_mil_reais)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

#### Save Dataset

```{r}
agricultural_gis <- consolidada %>%
  select(
    cod_ibge,
    sisagua_pct_deteccao_quartile,
    sisagua_n_pesticidas_detect_quartile,
    pam_area_colhida_ha_quartile,
    pam_valor_prod_mil_reais_quartile
  )


#save separate dataset per year

for (yr in 2014:2024) {
  consolidada %>%
    filter(ano == yr) %>%
    write_csv(paste0("agricultural_", yr, ".csv"))
}

```
