---
title: "SISAGUA — Pesticide Exposure in Drinking Water and Health Outcomes"
subtitle: "São Paulo state, 2014–2024 | Municipality × Year Panel"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

## 1. Setup

```{r}
library(arrow)
library(dplyr)
library(tidyr)
library(ggplot2)
library(skimr)
library(DataExplorer)
```

### 1.1 Directory setup
```{r}
# Set the directory so it works
knitr::opts_knit$set(root.dir = "c:/Users/alexa/Desktop/Banco de Dados - Intox. Agro") # set your path
```
---

## 2. Load raw data

```{r}
# Individual-level sources
sisagua_raw  <- read_parquet("resultados/SISAGUA/sisagua_sp_2014_2024.parquet")
sih_raw      <- read_parquet("resultados/SIH/sih_iexo_sp_2014_2024.parquet")
sinan_raw    <- read_parquet("resultados/SINAN/sinan_iexo_sp_2014_2024.parquet")
sim_raw      <- read_parquet("resultados/SIM/sim_iexo_sp_2014_2024.parquet")

# Population denominator (for rates)
pop_raw      <- read_parquet("resultados/contextual/populacao_sp_municipio_ano.parquet")
```

---

## 3. Build exposure: SISAGUA aggregated to municipality × year

Three detection thresholds are built in parallel so you can choose later.

```{r}
# Population denominator: total per municipality x year
pop <- pop_raw |>
  mutate(ano = as.integer(ano)) |>
  group_by(cod_ibge, ano) |>
  summarise(pop_total = sum(populacao, na.rm = TRUE), .groups = "drop")

# SISAGUA aggregation
sisagua <- sisagua_raw |>
  mutate(
    ano      = as.integer(NU_ANO),
    cod_ibge = as.character(cod_ibge)
  ) |>
  group_by(cod_ibge, ano) |>
  summarise(    # --- sample counts ---
    n_amostras            = n(),
    
    # --- threshold 1: NUMERICO only (quantifiable detections) ---
    n_detect_num          = sum(TIPO_RESULTADO == "NUMERICO", na.rm = TRUE),
    pct_detect_num        = 100 * n_detect_num / n_amostras,
    n_pesticidas_num      = n_distinct(PARAMETRO_FINAL[TIPO_RESULTADO == "NUMERICO"]),
    
    # --- threshold 2: NUMERICO + MENOR_LQ (detected at any level) ---
    n_detect_lq           = sum(TIPO_RESULTADO %in% c("NUMERICO", "MENOR_LQ"), na.rm = TRUE),
    pct_detect_lq         = 100 * n_detect_lq / n_amostras,
    
    # --- threshold 3: exceedances of regulatory limit (VMP) ---
    n_exceedances_vmp     = sum(
      TIPO_RESULTADO == "NUMERICO" &
      !is.na(RESULTADO_NUM) &
      !is.na(as.numeric(VMP)) &
      RESULTADO_NUM > as.numeric(VMP),
      na.rm = TRUE
    ),
    pct_exceedances_vmp   = 100 * n_exceedances_vmp / n_amostras,
    
    .groups = "drop"
  )
```

---

## 4. Build outcomes: SIH, SINAN, SIM aggregated to municipality × year

```{r}
# SIH — hospitalizations
sih <- sih_raw |>
  mutate(
    cod_ibge = as.character(MUNIC_RES),
    ano      = as.integer(ANO_CMPT)
  ) |>
  group_by(cod_ibge, ano) |>
  summarise(
    sih_n_hosp       = n(),
    sih_n_hosp_t60   = sum(substr(DIAG_PRINC, 1, 3) == "T60", na.rm = TRUE),
    sih_n_obitos     = sum(MORTE == "Sim", na.rm = TRUE),
    sih_dias_perm_media = mean(as.numeric(DIAS_PERM), na.rm = TRUE),
    .groups = "drop"
  )

# SINAN — poisoning notifications
sinan <- sinan_raw |>
  mutate(
    cod_ibge = as.character(ID_MN_RESI),
    ano      = as.integer(ano_origem)
  ) |>
  group_by(cod_ibge, ano) |>
  summarise(
    sinan_n_notif       = n(),
    sinan_n_notif_agric = sum(!is.na(LAVOURA) & LAVOURA != "", na.rm = TRUE),
    sinan_n_obitos      = sum(EVOLUCAO == "2", na.rm = TRUE),
    .groups = "drop"
  )

# SIM — deaths
sim <- sim_raw |>
  mutate(
    cod_ibge = as.character(CODMUNRES),
    ano      = as.integer(substr(DTOBITO, 1, 4))
  ) |>
  group_by(cod_ibge, ano) |>
  summarise(
    sim_n_obitos = n(),
    .groups = "drop"
  )
```

---

## 5. Join into a municipality × year panel

```{r}
# Full grid: all municipalities with population data x all years
panel <- pop |>
  left_join(sisagua, by = c("cod_ibge", "ano")) |>
  left_join(sih,     by = c("cod_ibge", "ano")) |>
  left_join(sinan,   by = c("cod_ibge", "ano")) |>
  left_join(sim,     by = c("cod_ibge", "ano")) |>
  # Replace NA counts with 0 (municipalities with no recorded events)
  mutate(across(c(starts_with("sih_n"), starts_with("sinan_n"), starts_with("sim_n")),
                \(x) replace_na(x, 0))) |>
  # Crude rates per 100,000
  mutate(
    taxa_hosp_100k        = 1e5 * sih_n_hosp   / pop_total,
    taxa_hosp_t60_100k    = 1e5 * sih_n_hosp_t60 / pop_total,
    taxa_notif_100k       = 1e5 * sinan_n_notif / pop_total,
    taxa_notif_agric_100k = 1e5 * sinan_n_notif_agric / pop_total,
    taxa_obitos_sim_100k  = 1e5 * sim_n_obitos  / pop_total
  )

glimpse(panel)
```

---

## 6. Data quality

### 6.1 Missingness

```{r}
plot_missing(panel)
```

> **Note:** `NA` in SISAGUA columns means the municipality had no water monitoring records
> that year — absence of monitoring, not absence of contamination.

### 6.2 Skimr summary

```{r}
skim(panel)
```

---

## 7. Exploratory analysis

### 7.1 Distribution of exposure variables (three thresholds)

```{r}
panel |>
  select(pct_detect_num, pct_detect_lq, pct_exceedances_vmp) |>
  plot_histogram()
```

### 7.2 Distribution of outcome rates

```{r}
panel |>
  select(taxa_hosp_100k, taxa_hosp_t60_100k,
         taxa_notif_100k, taxa_notif_agric_100k,
         taxa_obitos_sim_100k) |>
  plot_histogram()
```

### 7.3 Correlation between the three exposure thresholds

```{r}
panel |>
  select(pct_detect_num, pct_detect_lq, pct_exceedances_vmp) |>
  plot_correlation(type = "continuous")
```

### 7.4 Correlation between all exposure and outcome variables

```{r}
panel |>
  select(pct_detect_num, pct_detect_lq, pct_exceedances_vmp,
         taxa_hosp_100k, taxa_notif_100k, taxa_obitos_sim_100k) |>
  plot_correlation(type = "continuous")
```

---

## 8. Descriptive analysis

### 8.1 Annual mean of exposure thresholds

```{r}
panel |>
  group_by(ano) |>
  summarise(across(c(pct_detect_num, pct_detect_lq, pct_exceedances_vmp),
                   \(x) mean(x, na.rm = TRUE))) |>
  pivot_longer(-ano, names_to = "threshold", values_to = "pct_mean") |>
  ggplot(aes(ano, pct_mean, colour = threshold)) +
  geom_line(linewidth = 0.8) +
  geom_point() +
  labs(title = "Mean detection rate by threshold — SP municipalities",
       x = "Year", y = "Mean % of samples", colour = NULL) +
  theme_minimal()
```

### 8.2 Annual mean of outcome rates

```{r}
panel |>
  group_by(ano) |>
  summarise(across(c(taxa_hosp_100k, taxa_notif_100k, taxa_obitos_sim_100k),
                   \(x) mean(x, na.rm = TRUE))) |>
  pivot_longer(-ano, names_to = "outcome", values_to = "rate_mean") |>
  ggplot(aes(ano, rate_mean, colour = outcome)) +
  geom_line(linewidth = 0.8) +
  geom_point() +
  labs(title = "Mean outcome rates per 100k — SP municipalities",
       x = "Year", y = "Mean rate per 100,000", colour = NULL) +
  theme_minimal()
```

### 8.3 Scatter: SISAGUA detection vs. hospitalization rate

```{r}
panel |>
  filter(!is.na(pct_detect_num)) |>
  ggplot(aes(pct_detect_num, taxa_hosp_100k)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "firebrick") +
  facet_wrap(~ano) +
  labs(title = "Pesticide detection (NUMERICO) vs. hospitalization rate",
       x = "% samples with quantifiable detection",
       y = "Hospitalizations per 100,000") +
  theme_minimal()
```

### 8.4 Scatter: SISAGUA detection vs. notification rate

```{r}
panel |>
  filter(!is.na(pct_detect_num)) |>
  ggplot(aes(pct_detect_num, taxa_notif_100k)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "steelblue") +
  facet_wrap(~ano) +
  labs(title = "Pesticide detection (NUMERICO) vs. notification rate",
       x = "% samples with quantifiable detection",
       y = "Notifications per 100,000") +
  theme_minimal()
```

---

## 9. Analytical models

### 9.1 Linear model — hospitalization rate ~ detection rate

```{r}
fit_sih <- lm(taxa_hosp_100k ~ pct_detect_num + factor(ano),
              data = panel)
summary(fit_sih)
```

### 9.2 Linear model — notification rate ~ detection rate

```{r}
fit_sinan <- lm(taxa_notif_100k ~ pct_detect_num + factor(ano),
                data = panel)
summary(fit_sinan)
```

### 9.3 Linear model — mortality rate ~ detection rate

```{r}
fit_sim <- lm(taxa_obitos_sim_100k ~ pct_detect_num + factor(ano),
              data = panel)
summary(fit_sim)
```

---