---
title: "SIH"
output: html_document
date: "2026-06-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Packages

```{r}
library(DataExplorer)
library(arrow)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(knitr)
library(kableExtra)
library(stringr)
```

## Using Consolidated for Analyses

*This dataset already has transformed variables from SIH*

```{r}
consolidada <- read_parquet("~/GitHub/intox_agro/resultados/base_consolidada_sp_2014_2024.parquet")

#subset
poisoning <- consolidada |>
  select(cod_ibge,
         ano,
         sih_n_hosp_0_14,
         sinan_n_notif_0_14,
         sim_n_obitos_0_14,
         taxa_hosp_0_14_100k,
         taxa_notif_0_14_100k,
         taxa_obitos_sim_0_14_100k,
         taxa_letalidade_hosp_0_14,
         sih_n_obitos_hosp_0_14)
```

# Hospitalization per 100K

#### Missing

```{r}
plot_missing(poisoning[, c("taxa_hosp_0_14_100k")])
```

### Raw Frequency

```{r}
median(poisoning$taxa_hosp_0_14_100k)

mean(poisoning$taxa_hosp_0_14_100k)
```

#### Create Quartiles - same cuts across all years

```{r}

#zero heavy
summary(poisoning$taxa_hosp_0_14_100k)

quantile(
  poisoning$taxa_hosp_0_14_100k,
  probs = seq(0, 1, 0.25),
  na.rm = TRUE
)

#Calculate quartile breaks using only non-zero values
breaks_hosp <- poisoning %>%
  filter(taxa_hosp_0_14_100k > 0) %>%
  pull(taxa_hosp_0_14_100k) %>%
  quantile(probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)

# Create new variable in the full dataset
poisoning <- poisoning %>%
  mutate(
    taxa_hosp_0_14_100k_quartile = case_when(
      taxa_hosp_0_14_100k == 0 ~ "0",
      TRUE ~ as.character(
        cut(
          taxa_hosp_0_14_100k,
          breaks = breaks_hosp,
          include.lowest = TRUE,
          labels = c("1", "2", "3", "4")))))
```

## Table

```{r}

poisoning_summary <- poisoning %>%
  select(
    ano,
    taxa_hosp_0_14_100k,
    taxa_hosp_0_14_100k_quartile
  ) %>%
  rename(
    value = taxa_hosp_0_14_100k,
    Quartile = taxa_hosp_0_14_100k_quartile
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)


poisoning_summary

#kable
poisoning_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Quartile",
      "n",
      "Median Rate",
      "Range"
    ),
    caption = "Hospitalization Rate Quartiles by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)
```

## Plot

```{r}
#raw
hosp_summary <- consolidada %>%
  group_by(ano) %>%
  summarise(
    total_hosp = sum(sih_n_hosp_0_14, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_summary, aes(x = ano, y = total_hosp, group = 1)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(breaks = unique(hosp_summary$ano)) +
  labs(
    x = "Year",
    y = "Hospitalizations (n)",
    title = "Hospitalizations among children aged 0–14 years"
  ) +
  theme_minimal(base_size = 14)

#quartiles
ggplot(poisoning_summary,
       aes(x = ano, y = Median, color = factor(Quartile))) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = unique(sort(poisoning_summary$ano))
  ) +
  labs(
    title = "Median Hospitalization Rate Ages 0-14 years per 100K",
    x = "Year",
    y = "Median Hospitalization Rate per 100K",
    color = "Quartile"
  ) +
  theme_minimal()
```

## Ribbon Plot

```{r}

summary_quartile <- poisoning %>%
  group_by(ano, taxa_hosp_0_14_100k_quartile) %>%
  summarise(
    min = min(taxa_hosp_0_14_100k, na.rm = TRUE),
    median = median(taxa_hosp_0_14_100k, na.rm = TRUE),
    max = max(taxa_hosp_0_14_100k, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(summary_quartile,
       aes(x = ano,
           y = median,
           color = factor(taxa_hosp_0_14_100k_quartile),
           group = taxa_hosp_0_14_100k_quartile)) +
  geom_linerange(aes(ymin = min, ymax = max), alpha = 0.5) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(summary_quartile$ano)) +
  labs(
    title = "Median Hospitalization Rate Ages 0-14 years per 100K",
    x = "Year",
    y = "Hospitalizaitons per 100K",
    color = "Quartile"
  ) +
  theme_minimal()

ggplot(poisoning,
       aes(x = factor(ano),
           y = taxa_hosp_0_14_100k,
           fill = factor(taxa_hosp_0_14_100k_quartile))) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(
    x = "Year",
    y = "Hospitalizaitons per 100K",
    fill = "Quartile"
  ) +
  theme_minimal()
```

## Extreme Values

```{r}
range(poisoning$taxa_hosp_0_14_100k, na.rm = TRUE)

consolidada %>%
  filter(taxa_hosp_0_14_100k > 100) %>%
  arrange(desc(taxa_hosp_0_14_100k)) %>%
  select(cod_ibge, ano, taxa_hosp_0_14_100k)


#table
poisoning %>%
  group_by(cod_ibge) %>%
  summarise(
    n_years_over_100 = sum(taxa_hosp_0_14_100k > 100, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_years_over_100 > 0) %>%
  arrange(desc(n_years_over_100))
```

## Normality

```{r}
ggplot(poisoning, aes(x = taxa_hosp_0_14_100k)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

# Case-fatality Rate per 100K

#### Missing

```{r}
plot_missing(poisoning[, c("taxa_letalidade_hosp_0_14")])
```

### Raw Frequency

```{r}
median(poisoning$taxa_letalidade_hosp_0_14, na.rm = TRUE)

mean(poisoning$taxa_letalidade_hosp_0_14, na.rm = TRUE)
```

#### Create Quartiles

```{r}
#Calculate quartile breaks using only non-zero values
breaks_cfr <- poisoning %>%
  filter(taxa_letalidade_hosp_0_14 > 0) %>%
  pull(taxa_letalidade_hosp_0_14) %>%
  quantile(probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)

# Create new variable in the full dataset
poisoning <- poisoning %>%
  mutate(
    taxa_letalidade_hosp_0_14_q = case_when(
      taxa_sih_cfr == 0 ~ "0",
      TRUE ~ as.character(
        cut(
          taxa_letalidade_hosp_0_14,
          breaks = breaks,
          include.lowest = TRUE,
          labels = c("1", "2", "3", "4")))))
```

## Table

```{r}
fatality_summary <- poisoning %>%
  select(
    ano,
    taxa_letalidade_hosp_0_14,
    taxa_letalidade_hosp_0_14_q
  ) %>%
  rename(
    value = taxa_letalidade_hosp_0_14,
    Quartile = taxa_letalidade_hosp_0_14_q
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)


fatality_summary

#kable
fatality_summary %>%
  arrange(ano, Quartile) %>%
  kable(
    digits = 2,
    col.names = c(
      "Year",
      "Quartile",
      "n",
      "Median Rate",
      "Range"
    ),
    caption = "Case Fatality Rate Quartiles by Year"
  ) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(full_width = FALSE)

```

### Plot

```{r}
ggplot(fatality_summary,
       aes(x = ano, y = Median, color = factor(Quartile))) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = unique(sort(fatality_summary$ano))
  ) +
  theme_minimal()
```

## Normality

```{r}
ggplot(poisoning, aes(x = sih_cfr)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

## Raw Death Count

```{r}
death_summary <- poisoning %>%
  group_by(ano) %>%
  summarise(
    total_death = sum(sih_n_obitos_hosp_0_14, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(death_summary, aes(x = ano, y = total_death, group = 1)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(breaks = unique(death_summary$ano)) +
  labs(
    x = "Year",
    y = "Deaths (n)",
    title = "Deaths among Hospitalized Children 0–14 years"
  ) +
  theme_minimal(base_size = 14)

```

#### Save dataset for mapping

```{r}

for (yr in 2014:2024) {
  poisoning %>%
    filter(ano == yr) %>%
    write_csv(paste0("poisoning_", yr, ".csv"))
}
```

# Using SIH

```{r}
SIH <- read_parquet("~/GitHub/intox_agro/resultados/SIH/sih_iexo_sp_2014_2024.parquet")

```

```{r}
#clean

SIH_0_14 <- SIH |>
  filter(
    COD_IDADE %in% c("Dias", "Meses") |
      (COD_IDADE == "Anos" & IDADE >= 0 & IDADE <= 14))

SIH_0_14 %>%
  filter(str_sub(DIAG_PRINC, 1, 3) %in% c("X48", "X68", "X87", "Y18")) %>%
  count(DIAG_PRINC, sort = TRUE)

SIH_0_14 %>%
  filter(str_sub(DIAG_SECUN, 1, 3) %in% c("X48", "X68", "X87", "Y18")) %>%
  count(DIAG_SECUN, sort = TRUE)
```

### Types of Intoxication

```{r}

table(SIH_0_14$DIAG_PRINC)
```

### Age Groups

```{r}

```

### Race

```{r}

```

### Temporal Trends
