---
title: "SIH"
output: html_document
date: "2026-06-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### Packages

```{r}
library(DataExplorer)
library(arrow)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(knitr)
library(kableExtra)
library(stringr)
library(gt)
library(scales)
library(gtsummary)
library(flextable)
library(writexl)

```

## Using Consolidated for Analyses

*This dataset already has transformed variables from SIH*

```{r}
consolidada <- read_parquet("~/GitHub/intox_agro/resultados/base_consolidada_sp_2014_2024.parquet")

#subset
poisoning <- consolidada |>
  select(cod_ibge,
         ano,
         pop_total,
         sih_n_hosp_t60_0_14,
         taxa_hosp_t60_0_14_100k,
         sih_n_hosp_t60,
         sih_n_hosp_t60_0,
         sih_n_hosp_t60_1,
         sih_n_hosp_t60_2,
         sih_n_hosp_t60_3,
         sih_n_hosp_t60_4,
         sih_n_hosp_t60_8,
         sih_n_hosp_t60_9,
         )
```

# Hospitalization per 100K, ages 0-14 ICD codes T60

#### Missing

```{r}
plot_missing(poisoning)

```

### Aggregate Rate Frequencies

```{r}
median(poisoning$taxa_hosp_t60_0_14_100k)

mean(poisoning$taxa_hosp_t60_0_14_100k)
```

#### Create Quartiles - same cuts across all years

```{r}
#Quartile breaks using only non-zero values, same quartiles each year 
breaks_hosp <- poisoning %>%
  filter(taxa_hosp_t60_0_14_100k > 0) %>%
  pull(taxa_hosp_t60_0_14_100k) %>%
  quantile(probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)

# Create new variable in the full dataset
poisoning <- poisoning %>%
  mutate(
    taxa_hosp_t60_0_14_100k_q = case_when(
      taxa_hosp_t60_0_14_100k == 0 ~ "0",
      TRUE ~ as.character(
        cut(
          taxa_hosp_t60_0_14_100k,
          breaks = breaks_hosp,
          include.lowest = TRUE,
          labels = c("1", "2", "3", "4")))))

```

## Table

```{r}

hosp_quartile_summary <- poisoning %>%
  group_by(ano, taxa_hosp_t60_0_14_100k_q) %>%
  summarise(
    n_municipalities = n(),
    median_population = median(pop_total, na.rm = TRUE),
    median_hosp = median(taxa_hosp_t60_0_14_100k, na.rm = TRUE),
    min_pop = min(pop_total, na.rm = TRUE),
    max_pop = max(pop_total, na.rm = TRUE),
    min_hosp = min(taxa_hosp_t60_0_14_100k, na.rm = TRUE),
    max_hosp = max(taxa_hosp_t60_0_14_100k, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(taxa_hosp_t60_0_14_100k_q) %>%
  summarise(
    `Municipalities (median, range)` = paste0(
      round(median(n_municipalities)),
      " (",
      min(n_municipalities),
      "–",
      max(n_municipalities),
      ")"
    ),
    `Population (median, range)` = paste0(
      round(median(median_population)),
      " (",
      round(min(min_pop)),
      "–",
      round(max(max_pop)),
      ")"
    ),
    `Hospitalization rate (median, range)` = paste0(
      round(median(median_hosp), 2),
      " (",
      round(min(min_hosp), 2),
      "–",
      round(max(max_hosp), 2),
      ")"
    ),
    .groups = "drop"
  ) %>%
  rename(
    Quartile = taxa_hosp_t60_0_14_100k_q
  ) %>%
  arrange(Quartile)


hosp_quartile_summary

```

```{r}
#Above table for 2014, 2017, 2021, 2024
hosp_quartile_summary_years <- poisoning %>%
  filter(ano %in% c(2014, 2017, 2021, 2024)) %>%
  group_by(ano, taxa_hosp_t60_0_14_100k_q) %>%
  summarise(
    `Number of municipalities` = n(),
    `Population (median, range)` = paste0(
      round(median(pop_total, na.rm = TRUE)),
      " (",
      round(min(pop_total, na.rm = TRUE)),
      "–",
      round(max(pop_total, na.rm = TRUE)),
      ")"
    ),
    `Hospitalization rate (median, range)` = paste0(
      round(median(taxa_hosp_t60_0_14_100k, na.rm = TRUE), 2),
      " (",
      round(min(taxa_hosp_t60_0_14_100k, na.rm = TRUE), 2),
      "–",
      round(max(taxa_hosp_t60_0_14_100k, na.rm = TRUE), 2),
      ")"
    ),
    .groups = "drop"
  ) %>%
  rename(
    Year = ano,
    Quartile = taxa_hosp_t60_0_14_100k_q
  ) %>%
  arrange(Year, Quartile)

hosp_quartile_summary_years

# write_xlsx(hosp_quartile_summary_years, "hosp_quartile_summary_years.xlsx")

```

```{r}
#all year summary
hosp_rate_14_24_summary <- poisoning %>%
  select(
    ano,
    taxa_hosp_t60_0_14_100k,
    taxa_hosp_t60_0_14_100k_q
  ) %>%
  rename(
    value = taxa_hosp_t60_0_14_100k,
    Quartile = taxa_hosp_t60_0_14_100k_q
  ) %>%
  group_by(ano, Quartile) %>%
  summarise(
    n = n(),
    Median = median(value, na.rm = TRUE),
    Min = min(value, na.rm = TRUE),
    Max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Range = paste0(Min, "-", Max)
  ) %>%
  select(ano, Quartile, n, Median, Range)

hosp_rate_14_24_summary
```

## Plot

```{r}
#raw
hosp_summary <- poisoning %>%
  group_by(ano) %>%
  summarise(
    total_hosp = sum(sih_n_hosp_t60_0_14, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(hosp_summary,
       aes(x = ano, y = total_hosp)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 2014:2024) +
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  labs(
        title = "Number of T60 Hospitalizations ages 0-14 years",

    x = "Year",
    y = "Hospitalizations (n)"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") 

#rate plot
rate_summary <- poisoning %>%
  group_by(ano) %>%
  summarise(
    hosp_rate = sum(taxa_hosp_t60_0_14_100k, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(rate_summary, aes(x = ano, y = hosp_rate, group = 1)) +
   geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 2014:2024) +
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  labs(
    x = "Year",
    y = "Hospitalizations per 100K",
    title = "T60 Hospitalization Rate Among children aged 0–14 years"
  ) +
theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") 
#quartiles

ggplot(hosp_rate_14_24_summary,
       aes(x = ano, y = Median, color = factor(Quartile))) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = unique(sort(hosp_rate_14_24_summary$ano))
  ) +
  labs(
    title = "Median T60 Hospitalization Rate Ages 0-14 years per 100K",
    x = "Year",
    y = "Median Hospitalization Rate per 100K",
    color = "Quartile"
  ) +
  theme_minimal()
```

## Ribbon Plot

```{r}

summary_quartile <- poisoning %>%
  group_by(ano, taxa_hosp_t60_0_14_100k_q) %>%
  summarise(
    min = min(taxa_hosp_t60_0_14_100k, na.rm = TRUE),
    median = median(taxa_hosp_t60_0_14_100k, na.rm = TRUE),
    max = max(taxa_hosp_t60_0_14_100k, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  summary_quartile,
  aes(
    x = ano,
    y = median,
    color = factor(taxa_hosp_t60_0_14_100k_q),
    group = taxa_hosp_t60_0_14_100k_q
  )
) +
  geom_linerange(aes(ymin = min, ymax = max), alpha = 0.5) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(summary_quartile$ano)) +
  labs(
    title = "Median T60 Hospitalization Rate Ages 0–14 years per 100K",
    x = "Year",
    y = "Median Hospitalizations per 100K",
    color = "Quartile"
  ) +
  theme_minimal()
  
```

## Extreme Values

```{r}
range(poisoning$taxa_hosp_t60_0_14_100k, na.rm = TRUE)

consolidada %>%
  filter(taxa_hosp_t60_0_14_100k > 30) %>%
  arrange(desc(taxa_hosp_t60_0_14_100k)) %>%
  select(cod_ibge, ano, taxa_hosp_t60_0_14_100k)


#table
poisoning %>%
  group_by(cod_ibge) %>%
  summarise(
    n_years_over_30 = sum(taxa_hosp_t60_0_14_100k > 30, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_years_over_30 > 0) %>%
  arrange(desc(n_years_over_30))
```

## Normality

```{r}
ggplot(poisoning, aes(x = taxa_hosp_t60_0_14_100k)) +
  geom_histogram(bins = 30) +
  theme_minimal()
```

#### Save dataset for mapping

```{r}

#for (yr in 2014:2024) {
#  poisoning %>%
#    filter(ano == yr) %>%
#    write_csv(paste0("poisoning_", yr, ".csv"))}
```

# Using SIH

```{r}
SIH <- read_parquet("~/GitHub/intox_agro/resultados/SIH/sih_iexo_sp_2014_2024.parquet")

```

```{r}
#Pesticides Only
SIH_0_14_T60 <- SIH |>
  filter(substr(DIAG_PRINC, 1, 3) == "T60") |>
  filter(
    COD_IDADE %in% c("Dias", "Meses") |
      (COD_IDADE == "Anos" & as.integer(IDADE) <= 14))


unique(SIH_0_14_T60$DIAG_PRINC)
```

## Missingness

```{r}
plot_missing(SIH_0_14_T60)

sum(is.na(SIH_0_14_T60$RACA_COR))

SIH_0_14_T60 %>%
  filter(is.na(RACA_COR)) %>%
  distinct(MUNIC_RES)

```

## Descriptive Table of Municipalities

```{r}
#Municipality dataset

pop_0_14_and_rates_by_municipality_2014_2024 <- read_csv("~/GitHub/intox_agro/resultados/pop_0_14_and_rates_by_municipality_2014_2024.csv")

```

```{r}
SIH_0_14_count <- SIH_0_14_T60 %>%
  count(MUNIC_RES) %>%
  arrange(MUNIC_RES) %>%
  mutate(
    `n (%)` = paste0(n, " (", percent(n / sum(n), accuracy = 0.1), ")")
  ) %>%
  select(MUNIC_RES, `n (%)`)

SIH_0_14_count

#join
SIH_0_14_count <- SIH_0_14_count %>%
  mutate(MUNIC_RES = as.numeric(MUNIC_RES))

munic_cases <- pop_0_14_and_rates_by_municipality_2014_2024 %>%
  semi_join(SIH_0_14_count, by = c("cod_ibge" = "MUNIC_RES"))

#just rates
munic_rates <- munic_cases %>%
  select(cod_ibge, hosp_pest_rate_0_14_per100k_yr) %>%
  arrange(cod_ibge)

munic_rates

#add range from consolidated
SIH_0_14_count <- SIH_0_14_count %>%
  mutate(MUNIC_RES = as.character(MUNIC_RES))

consolidada_filtered <- consolidada %>%
  select(cod_ibge, ano, taxa_hosp_t60_0_14_100k, pop_0_14) %>%
  semi_join(
    SIH_0_14_count,
    by = c("cod_ibge" = "MUNIC_RES")
  )

munic_rate_range <- consolidada_filtered %>%
  group_by(cod_ibge) %>%
  summarise(
    `Median hospitalization rate` = round(
      median(taxa_hosp_t60_0_14_100k, na.rm = TRUE), 1
    ),
    `Hospitalization rate range` = paste0(
      round(min(taxa_hosp_t60_0_14_100k, na.rm = TRUE), 1),
      "–",
      round(max(taxa_hosp_t60_0_14_100k, na.rm = TRUE), 1)
    ),
    .groups = "drop"
  )


munic_rate_range
```

## Types of Intoxication

```{r}
#table
diag_table <- SIH_0_14_T60 %>%
  count(DIAG_PRINC, name = "Frequency") %>%
  mutate(
    Percent = Frequency / sum(Frequency)) %>%
  arrange(desc(Frequency))

diag_table %>%
  gt() %>%
  fmt_percent(
    columns = Percent,
    decimals = 1) %>%
  cols_label(
    DIAG_PRINC = "Principal Diagnosis (ICD-10 Code)",
    Frequency = "n",
    Percent = "%") %>%
  tab_header(
    title = md("**Distribution of Principal Diagnoses Among Hospitalizations of Children Aged 0–14 Years**")) %>%
  fmt_number(
    columns = Frequency,
    decimals = 0,
    use_seps = TRUE)

#plot
ggplot(SIH_0_14_T60, aes(x = DIAG_PRINC)) +
  geom_bar(fill = "#2171B5", color = "black") +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +
  xlab("Principal Diagnosis") +
  ylab("Count")

```

#### By Year

```{r}
diag_plot <- SIH_0_14_T60 %>%
  count(ANO_CMPT, DIAG_PRINC, name = "Frequency") %>%
  group_by(ANO_CMPT) %>%
  slice_max(Frequency, n = 10, with_ties = FALSE) %>%
  ungroup()

ggplot(diag_plot,
       aes(x = factor(ANO_CMPT),
           y = DIAG_PRINC,
           size = Frequency)) +
  geom_point(color = "#2171B5") +
  scale_size(range = c(3, 10)) +
  labs(
    x = "Year",
    y = "ICD-10 code",
    size = "Hospitalizations"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") 
```

#### By Age Group

```{r}
#Create var
SIH_0_14_T60 <- SIH_0_14_T60 %>%
  mutate(
    IDADE = as.numeric(IDADE),
    age_group = case_when(
      COD_IDADE %in% c("Dias", "Meses") ~ "0–3 years",
      COD_IDADE == "Anos" & IDADE >= 0 & IDADE <= 3 ~ "0–3 years",
      COD_IDADE == "Anos" & IDADE >= 4 & IDADE <= 7 ~ "4–7 years",
      COD_IDADE == "Anos" & IDADE >= 8 & IDADE <= 10 ~ "8–10 years",
      COD_IDADE == "Anos" & IDADE >= 11 & IDADE <= 14 ~ "11–14 years",
      TRUE ~ NA_character_),
    age_group = factor(
      age_group,
      levels = c("0–3 years", "4–7 years", "8–10 years", "11–14 years")))
```

```{r}
#table

#plot
diag_age <- SIH_0_14_T60 %>%
  count(age_group, DIAG_PRINC)

ggplot(diag_age,
       aes(x = age_group, y = n, fill = DIAG_PRINC)) +
  geom_col(position = "fill") +
scale_fill_manual(values = palette.colors(10, palette = "Tableau 10")) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1)
  )+
  theme_classic(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +
  labs(
    x = "Age group",
    y = "Hospitalizations (%)",
    fill = "Principal Diagnosis")
```

#### By Sex

```{r}
#plot
diag_sex <- SIH_0_14_T60 %>%
  count(SEXO, DIAG_PRINC)

ggplot(diag_sex,
       aes(x = SEXO, y = n, fill = DIAG_PRINC)) +
  geom_col(position = "fill") +
scale_fill_manual(values = palette.colors(10, palette = "Tableau 10")) +
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +
  labs(
    x = "Sex",
    y = "Hospitalizations (%)",
    fill = "Principal Diagnosis")
```

#### By Race

```{r}
#Recode
SIH_0_14_T60 <- SIH_0_14_T60 %>%
  mutate(
    RACA_COR = case_when(
      RACA_COR == "01" ~ "Branca (White)",
      RACA_COR == "02" ~ "Preta (Black)",
      RACA_COR == "03" ~ "Parda (Brown/Mixed)",
      RACA_COR == "04" ~ "Amarela (Asian)",
      RACA_COR == "05" ~ "Indígena (Indigenous)",
      is.na(RACA_COR) ~ "Missing",
      TRUE ~ "Unknown"
    )
  )
```

```{r}
#Stacked
diag_age <- SIH_0_14_T60 %>%
  count(RACA_COR, DIAG_PRINC)

ggplot(diag_age,
       aes(x = RACA_COR, y = n, fill = DIAG_PRINC)) +
  geom_col(position = "fill") +
scale_fill_manual(values = palette.colors(10, palette = "Tableau 10")) +
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +
  labs(
    x = "Race",
    y = "Hospitalizations (%)",
    fill = "Principal diagnosis")
```

## Age Groups

```{r}
#Table
age_table <- SIH_0_14_T60 %>%
  count(age_group, name = "Frequency") %>%
  mutate(
    Percent = Frequency / sum(Frequency)) %>%
  arrange(desc(Frequency))

age_table %>%
  gt() %>%
  fmt_percent(
    columns = Percent,
    decimals = 1) %>%
  cols_label(
    age_group = "Age Group",
    Frequency = "n",
    Percent = "%") %>%
  tab_header(
    title = md("**Distribution of Ages Amongst Hospitalizations**")) %>%
  fmt_number(
    columns = Frequency,
    decimals = 0,
    use_seps = TRUE)

#Aggregate Plot
ggplot(SIH_0_14_T60, aes(x = age_group, fill = age_group)) +
  geom_bar(aes(y = after_stat(count / sum(count)))) +
  geom_text(
    stat = "count",
    aes(
      y = after_stat(count / sum(count)),
      label = scales::percent(after_stat(count / sum(count)), accuracy = 0.1)
    ),
    vjust = -0.3) +
  scale_fill_brewer(palette = "Blues")+
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +
  labs(
    x = "Age Group",
    y = "Hospitalizations (%)",
    fill = "Age Group")

#Trend Plot
ggplot(SIH_0_14_T60,
       aes(x = ANO_CMPT, fill = age_group)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Blues")+
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +
  labs(
    x = "Year",
    y = "Hospitalizations (%)",
    fill = "Age Group")
```

### Race

```{r}
#Aggregate Plot
ggplot(SIH_0_14_T60, aes(x = RACA_COR, fill = RACA_COR)) +
  geom_bar(aes(y = after_stat(count / sum(count)))) +
  geom_text(
    stat = "count",
    aes(
      y = after_stat(count / sum(count)),
      label = scales::percent(after_stat(count / sum(count)), accuracy = 0.1)),
    hjust = -0.2) +
  coord_flip() +
scale_fill_manual(values = palette.colors(10, palette = "Tableau 10")) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.1))) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +  
  labs(
    x = "Race",
    y = "Percentage of Hospitalizations",
    fill = "Race") +
  theme(
    legend.position = "none")
```

### Sex

```{r}
#Table
sex_table <- SIH_0_14_T60 %>%
  count(SEXO, name = "Frequency") %>%
  mutate(
    Percent = Frequency / sum(Frequency)) %>%
  arrange(desc(Frequency))

sex_table %>%
  gt() %>%
  fmt_percent(
    columns = Percent,
    decimals = 1) %>%
  cols_label(
    SEXO = "Sex",
    Frequency = "n",
    Percent = "%") %>%
  tab_header(
    title = md("**Distribution of Hospitalizations by Sex**")) %>%
  fmt_number(
    columns = Frequency,
    decimals = 0,
    use_seps = TRUE)

#Aggregate Plot
ggplot(SIH_0_14_T60, aes(x = SEXO, fill = SEXO)) +
  geom_bar(aes(y = after_stat(count / sum(count)))) +
  geom_text(
    stat = "count",
    aes(
      y = after_stat(count / sum(count)),
      label = scales::percent(after_stat(count / sum(count)), accuracy = 0.1)
    ),
    vjust = -0.3) +
scale_fill_manual(values = palette.colors(10, palette = "Tableau 10")) +
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +
  labs(
    x = "Sex",
    y = "Hospitalizations (%)",
    fill = "Sex")

#Trend Plot
#Trend Plot
ggplot(SIH_0_14_T60,
       aes(x = ANO_CMPT, fill = SEXO)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Dark2")+
  scale_y_continuous(labels = scales::percent) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "right") +
  labs(
    x = "Year",
    y = "Hospitalizations (%)",
    fill = "Sex")
```

## Days of Stay

```{r}

SIH_0_14_T60 <- SIH_0_14_T60 %>%
  mutate(DIAS_PERM = as.numeric(DIAS_PERM))

```

#### By Race

```{r}
dias_race <- SIH_0_14_T60 %>%
  group_by(RACA_COR) %>%
  summarise(
    n = n(),
    `Median (Range)` = paste0(
      median(DIAS_PERM, na.rm = TRUE),
      " (",
      min(DIAS_PERM, na.rm = TRUE),
      "–",
      max(DIAS_PERM, na.rm = TRUE),
      ")"
    ),
    .groups = "drop"
  )

dias_race
```

#### By Sex

```{r}
dias_sex <- SIH_0_14_T60 %>%
  group_by(SEXO) %>%
  summarise(
    n = n(),
    `Median (Range)` = paste0(
      median(DIAS_PERM, na.rm = TRUE),
      " (",
      min(DIAS_PERM, na.rm = TRUE),
      "–",
      max(DIAS_PERM, na.rm = TRUE),
      ")"
    ),
    .groups = "drop"
  )

dias_sex
```

#### By Age Group

```{r}
dias_age <- SIH_0_14_T60 %>%
  group_by(age_group) %>%
  summarise(
    n = n(),
    `Median (Range)` = paste0(
      median(DIAS_PERM, na.rm = TRUE),
      " (",
      min(DIAS_PERM, na.rm = TRUE),
      "–",
      max(DIAS_PERM, na.rm = TRUE),
      ")"
    ),
    .groups = "drop"
  )

dias_age
```

## Table 1

```{r}

table1SIH <- SIH_0_14_T60 %>%
  select(
    SEXO,
    RACA_COR,
    age_group,
    DIAG_PRINC
  ) %>%
  tbl_summary(
    label = list(
      SEXO ~ "Sex",
      RACA_COR ~ "Race",
      age_group ~ "Age group",
      DIAG_PRINC ~ "Primary diagnosis (ICD-10)"
    ),
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    percent = "column",
    missing = "ifany",
    missing_text = "Missing"
  ) %>%
  bold_labels()

table1SIH


```
